---
title: "European ESG Performance and Banking Risk"
author: "Maurits de Haan (s8161658, m.c.p.dehaan@students.uu.nl)"
date: | 
  | 3rd of April, 2025
  | Individual internship for FNO Zorg voor Kansen
  | Course: Data Science Lab (ECB3DSL), Academic year 2024/2025
  | Supervisor: Dr. Ulrich Zierahn-Weilage
  | Word Count: XXXX (max 5000)
output:
  bookdown::pdf_document2:
      toc: false
      toc_depth: 3
      number_sections: true
      fig_height: 4
      fig_width: 6
      includes:
        in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)
```

```{r}
# Load required libraries
library(plm)
library(dplyr)
library(stargazer)
library(kableExtra)
library(modelsummary)
library(broom)
library(ggplot2)
library(gridExtra)
library(plm)
library(AER)
library(gridExtra)
library(kableExtra)
library(modelsummary)
library(plm)
library(dplyr)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(patchwork)
library(modelsummary)
library(broom)
library(knitr)
library(kableExtra)

sink(file = tempfile())
invisible(source(here::here("R", "import_ESG.R")))
invisible(source(here::here("R", "import_financial.R")))
invisible(source(here::here("R", "merging_ESG_financial.R")))
invisible(source(here::here("R", "analysis_fixed_effects_quadratic.R")))
invisible(source(here::here("R", "analysis_quantile.R")))
invisible(source(here::here("R", "analysis_pca.R")))
invisible(source(here::here("R", "analysis_iv.R")))
invisible(source(here::here("R", "analysis_exp.R")))
sink()
```

# Abstract {.unnumbered}

This study examines the relationship between ESG performance and risk in European banks (2020-2023) using EBA regulatory data and LSEG Refinitiv ESG scores. Through panel data analysis with fixed effects, quantile regression, and principal component analysis, several findings emerge. While overall ESG scores show limited association with risk metrics, the social pillar proves most influential, exhibiting significant positive relationships with capital measures in inverted U-shaped patterns. ESG effects vary across the risk distribution, with strongest impact in lower to middle quantiles. PCA identifies ESG structures beyond traditional categorizations, revealing strategic sustainability approaches with different risk implications. These findings inform both risk management and regulatory frameworks.

\newpage

| Statements |
|------------------------------|
| **Statement of Originality** | 
| I, the aforementioned student, herewith declare to have written this document and that I am responsible for the content of it. I declare that the text and the work presented in this document is original and that no sources other than those mentioned in the text and its references have been used in creating it. |
| Utrecht University School of Economics is responsible solely for the supervision of completion of the work, not for the content. |
| All questions about this paper can be directed to the author. | 
|------------------------------|
| **Disclosure Statement** |
| In this project, I have made use of the following Generative AI tools: ChatGPT 4o (reasoning), Claude 3.5 Sonnet (coding) |
| ChatGPT 4o was used to help with reasoning about the application of the methodology that would be usefull to use and the way in which this can be implemented. It was used as an extension of my own knowledge and as an external check on the assumptions I made about certain methodologies. Claude 3.5 Sonnet was utilised for certain coding tasks, mostly when it was outside of the scope of what I have learned earlier. All code that was provided by the model was thouroughly checked and validated if it was actully true (and in most cases I had to rewrite it completely). Next to that, Claude was used to integrate some functions in the code that I wrote. Claude was also used to help with for example a .tex and .css file to include the image in the header of this file and other "nice-to-have" things. |
| All content in this report was written by myself and not by a GenAI tool. Most, was first written by hand and then typed in RStudio. | 
|------------------------------|
| **Signature**
| |
| |
| |

\newpage 

\tableofcontents

\newpage

# Introduction

\begin{wrapfigure}{r}{0.5\textwidth}
  \centering
  \includegraphics[width=0.48\textwidth]{esg_structure.png}
  \caption{ESG Structure with Pillars and Components}
  \label{fig:esg-structure}
\end{wrapfigure}

In recent years, the role of environmental, social, and governance (ESG)  considerations in banking has gained significant attention in both academic and policy fields. Growing concerns about climate change, inequality and good governance have increased the interest how corporations and banks through their lending decision can drive the transition to a more sustainable and equitable economy (Carnevale & Drago, 2024; European Banking Authority, 2018; EU, 2014). Many banks across Europe are subject to Pillar 3 disclosures under the Capital Requirements Regulation (CRR), and new ESG disclosures further highlight climate and sustainability exposures (De Barros Fritz & Gentile, 2024). Policymakers and financial experts argue that improved ESG performance may help banks reduce their overall risk exposure, particularly by enhancing their reputation, improving stakeholder trust, and anticipating regulatory changes. However, whether ESG practices truly mitigate banking risk or whether lower-risk institutions merely find it easier to allocate resources toward socially responsible projects remains an open empirical question. This paper contributes to this debate by focusing on capital adequacy which reflects banking risk and using regulatory data instead of annual reports. In this paper, the relationship between ESG performance and capital ratios for European banks from 2020 to 2023 is analysed.

From a societal perspective, understanding whether strong ESG engagement leads to more resilient financial institutions carries high relevance. If banks with robust ESG strategies exhibit lower risk, this supports integrating ESG considerations into prudential regulation and highlights banks’ capacity to contribute to broader societal objectives without jeopardizing their stability. Conversely, if safer banks are simply better able to “afford” ESG investments, this suggests a different policy focus (e.g. financial health determines ESG commitments needed to be made). From a scientific standpoint, this study adds nuance to a growing literature that often uses conventional balance-sheet data to proxy risk, which may not fully capture regulatory metrics. By leveraging publicly available data from the European Banking Authority (EBA), a more reliable view of risk-weighted assets, Tier 1 capital, leverage, and provision levels is gained. Pairing these risk metrics with granular ESG data from LSEG Refinitiv further strengthens the empirical inquiry, because the ESG scores (and their component pillars) benefit from refined post-2019 methodologies.

This paper employs complementary methodological approaches to address the relationship between ESG performance and banking risk. Leveraging panel data techniques with two-way fixed effects, I account for unobserved heterogeneity while exploring potential non-linear relationships through quadratic specifications. To examine the heterogeneity of effects across different risk profiles, I implement quantile regression analysis, which reveals how ESG factors influence banks at various points in the risk distribution. Additionally, Principal Component Analysis (PCA) is applied to address multicollinearity concerns and uncover the underlying structure of ESG dimensions beyond the traditional pillar categorization.

The empirical results reveal several important insights. First, the overall ESG score shows a more nuanced relationship with risk metrics than previously documented in the literature. Second, disaggregating ESG into its constituent pillars uncovers substantial heterogeneity in their effects, with the social pillar demonstrating the strongest and most consistent association with capital adequacy measures. Third, these relationships vary significantly across the risk distribution, with more pronounced effects for banks in the middle quantiles. Finally, the PCA results suggest that distinct underlying dimensions of ESG performance exist beyond the conventional environmental, social, and governance categorization, each exhibiting different relationships with banking risk measures.

The remainder of this paper is structured as follows. Section 2 reviews the relevant literature on ESG performance and banking risk. Section 3 describes the data sources and variable construction. Section 4 provides descriptive statistics and preliminary analyses. Section 5 outlines the methodological approaches in detail. Section 6 presents and discusses the empirical results. Sections 7 concludes. 

# Literature Review

The growing incorporation of environmental, social, and governance (ESG) considerations in banking decision-making has led to a growing body of research analysing whether and how ESG practices shape financial risk and performance. 

## ESG and firm capital

First, the fundamental papers look at the corporate side, demonstrating that firms with stronger ESG profiles benefit from lower cost of debt and better credit ratings. For instance, Apergis et al. (2022) report that lower ESG ratings tend to increase the default risk premium charged by debtholders, thus suggesting that ESG factors influence the pricing of unsecured corporate debt.

Further connections between ESG metrics and cost of capital have been explored by Chava (2014), who finds that firms who are involved in many environmental controvocies are penalized with higher equity and debt costs. While the magnitude of this penalty varies across subcomponents of ESG, it does suggest that, overall, markets incorporate sustainability signals into the required returns on capital. Yet, other scholars remain more cautious about conclusive results. Menz (2010), studying European corporate bonds, finds no strong link between corporate social responsibility (CSR) and secondary market credit spreads, an outcome possibly attributed to the incomplete integration of ESG in bond pricing at the time. Similarly, Amiraslani et al. (2017) observe that the positive effect of strong social capital emerges primarily in times of crisis, such as during the 2008–2009 financial turmoil, when trust and default risk dominate market concerns. They show that high-social-capital firms could secure funding at lower costs during the height of the crisis but found no pronounced link between social capital and bond spreads in more stable years.

Work on the relationship between ESG factors and bank loan pricing has grown. Goss and Roberts (2010) detail how banks acquire additional information from corporate ESG ratings, which translates to higher loan spreads on firms with CSR concerns. Banks effectively penalize borrowers with shortcoming ESG profiles unless those loans are well secured. Subsequent reviews corroborate that higher ESG risk can lead to elevated bank lending costs (Carnevale & Drago, 2024). However, while evidence of this ESG “penalty” is common, differences often arise across ESG pillars, with governance sometimes playing a more substantial role in lowering perceived borrower risk.

## ESG and lending by banks

A parallel literature narrows its scope to banks themselves, examining whether ESG performance among banking institutions correlates with risk proxies such as default risk, nonperforming loans, tail risk, and market-based measures of capital adequacy. Curcio et al. (2024), for instance, use a sample of European banks and insurers to show that stronger ESG performance in nonfinancial counterparties reduces banks’ overall systemic risk exposure, albeit with nuanced differences when environmental and social facets are considered separately. 

Empirical efforts to measure direct effects of ESG on bank-specific outcomes often rely on risk-weighted capital ratios or default probability metrics. Several papers employ Merton-style distance to default or Z-scores to capture how ESG performance might reduce a bank’s vulnerability to extreme negative shocks (Chiaramonte et al., 2022; Danisman & Tarazi, 2023). The emerging consensus suggests that ESG considerations relate meaningfully to bank resilience. However, it remains unclear how the pathways are caused (e.g. via lower loan losses, improved reputational capital, or heightened stakeholder trust) (Gehrig et al., 2023). There is also extensive research focusing on operational risk: Galletta et al. (2022) document that higher ESG scores consistently reduce banks’ operational loss events, reflecting lower reputational and compliance risk. 

Despite a shared recognition that sustainability dimensions matter for financial risk of banks, important theoretical and empirical debates persist if this relationship can be identified as causal. One strand, often related to stakeholder theory, proposes that deliberate ESG initiatives improve a bank’s reputation, stakeholder satisfaction, and regulatory preparedness (Godfrey, 2005; He et al., 2021). Consequently, these banks have stronger buffers against unforeseen market stresses, which may translate into stronger capital ratios or lower cost of funding. Another line of inquiry posits that less risky institutions simply have the resources and latitude to engage in more robust ESG activities (Bouslah et al., 2013; Cornett et al., 2016). In that scenario, risk is arguably a precursor, rather than an outcome, of sustainability engagement. Studies such as Gehrig et al. (2023) and Danisman and Tarazi (2023) attempt to mitigate such endogeneity and simultaneity concerns by using instrumental variables or dynamic panel methodologies. 

## ESG Pillar scores and lending by banks

Researchers have further highlighted that different ESG pillars do not necessarily exert uniform effects on a bank’s capital adequacy. The governance dimension, for instance, has repeatedly been identified as a strong predictor of stability because it reflects managerial skill, board structure, and the capacity to address emerging sustainability concerns systematically (Chodnicka-Jaworska, 2021; Cohen, 2023). Environmental scores may be particularly relevant for banks exposed to climate-sensitive sectors. For example, some studies show minimal or inconsistent links between social scores and measures such as credit spreads or default risks (Amiraslani et al., 2017; Izcan & Bektas, 2022), whereas others find that employee relations and community engagements can bolster goodwill and mitigate downside risk in stress periods (Ashwin Kumar et al., 2016).

# Data 

The dataset employed in this study combines two sources of bank-level information for EBA member countries: regulatory bank data and ESG data. This paper focusses on European banks that are a member of the regulatory institution the European Banking Authority (EBA). The member countries include all Eurozone counties, apart from Croatia and Slovakia, non-Eurozone countries that are transitioning to adopting the euro (Bulgaria, Hungary, Poland, Romania and Sweden) and Denmark (opt-out for Eurozone participation). Next to that, the EBA also has some non-EU countries as members, such as Norway, Iceland and Liechtenstein. These countries are all included in the initial data-gathering. For all countries, the included banks are those with their headquarter in that country. 

## Regulatory bank data

For the regulatory financial data, the EBA data from the EU-Wide Transparency Exercise is used. This is validated date by banks and EU supervisors  (regarded to as "Historical KM"). The data is available from the third quarter of 2019 to the third quarter of 2024. Since the ESG data is in full years, only the data from 2020 to 2023 is used and the yearly values are calculated. This data contains standardized regulatory metrics for Eurozone banks, including, among others, total assets, total equity, loan and advance amounts, provisions, and various capital ratios. Originally the data contained over 12,000 observations, since each bank's financial metric was recorded as a separate observation. The data was therefore pivoted wider, so that a more structures and easier to merge dataset originates. Each financial variable now has its own column. Next to that, several financial ratios were calculated, for example the equity to assets, loan to assets and provision ratio. All variables in the dataset can be found in \@ref(tab:variables-table) in the appendix. 

## ESG data

The Refinitiv database of the London Stock Exchange Group (LSEG) was used for collecting the ESG ratings. Refinitiv is the most widely used database in academic and operational analysis about sustainability ratings. ESG ratings and reporting practices have undergone refinement in recent years, especially after 2019 (Mandas et al., 2024; Fitch, 2022; Moody’s, 2022; S&P Global, 2022), implying that more recent data are likely to capture richer and more standardized information about sustainability. Papers examining European banks from 2020 onward can thus rely on these improved ESG data sets (Curcio et al., 2024; Sun et al., 2024). Such refinements are crucial because risk-related metrics, especially regarding capital buffers, hinge on regulatory definitions and supervisory guidelines, which have also evolved to incorporate ESG-related climate stress testing (BCBS, 2020; EBA, 2021).

During the data retrieving process, first all public and private companies were filtered on ICB sector code (Banking, 301010) and then on the countries where these banks have their headquarters corresponding to the EBA financial dataset. Then, the ESG variables were selected. This was done broadly, so the general ESG scores, pillar scores and component scores were all included. This namely makes it possible to execute principal component analysis later. The ESG scores were retrieved as scores (range 1-100, higher is better) rather than letter-grade ratings. This was done to make quantitative analysis more precise and understandable. Also, the NULL variables were changes to NA, the year columns were changes to rows (so that each bank in a certain year has their own row) and variable names were uniformed. 

## Merging 

The merged dataset was constructed by performing an inner join between the regulatory data (EBA) and ESG data (Refinitiv). This merging approach ensured that only banks and years appearing in both datasets are retained for the final analytical sample. This ensures that each record in the final dataset contains both the regulatory capital and risk information and the ESG performance indicators for a given institution-year observation. Since the bank names differ between the two datasets, the Legal Entity Identifier (LEI) was used to merge the datasets. 

# Descriptive Analysis

This section provides a focused examination of key patterns in the data that inform tge understanding of the ESG-risk relationship in European banking. The sample comprises 132 bank-year observations across 2020-2023.

```{r esg-distribution, fig.cap="Distribution of ESG Scores and Pillar Components"}
# Create a more informative visualization showing both the overall distribution and pillar breakdown
p1 <- ggplot(data_analysis, aes(x=esg_score)) + 
  geom_histogram(bins=15, fill="#4472C4", alpha=0.7) +
  geom_vline(aes(xintercept=mean(esg_score, na.rm=TRUE)), 
             color="red", linetype="dashed", size=1) +
  labs(x="ESG Score", y="Frequency") +
  theme_minimal()

p2 <- data_analysis %>%
  select(environmental_pillar_score, social_pillar_score, governance_pillar_score) %>%
  pivot_longer(cols=everything(), names_to="Pillar", values_to="Score") %>%
  mutate(Pillar = factor(Pillar, 
                         levels=c("environmental_pillar_score", 
                                  "social_pillar_score", 
                                  "governance_pillar_score"),
                         labels=c("Environmental", "Social", "Governance"))) %>%
  ggplot(aes(x=Pillar, y=Score, fill=Pillar)) +
  geom_boxplot(alpha=0.7) +
  labs(x="", y="Score") +
  scale_fill_manual(values=c("#4472C4", "#ED7D31", "#A5A5A5")) +
  theme_minimal() +
  theme(legend.position="none")

grid.arrange(p1, p2, ncol=2)
```

Figure \@ref(fig:esg-distribution) reveals that ESG scores among European banks cluster around a mean of 73.32 (on a scale of 0-100), with a standard deviation of 11.95. The right panel demonstrates the distinct performance patterns across ESG pillars: banks score highest on environmental metrics (mean = 85.98), followed by social (75.36) and governance (68.62). This environmental prioritization likely reflects the European banking sector's response to recent EU policy initiatives such as the European Green Deal and Taxonomy Regulation.

Regarding risk-weighted capital measures, banks maintain an average CET1 risk exposure ratio of 0.60, with Tier 1 and total capital risk exposures at 0.67 and 0.76 respectively. Table 1 reveals that larger banks (Q4) tend to maintain higher capital ratios than smaller institutions (Q1), suggesting different capital management strategies across the size spectrum. The provisions ratio (mean = 0.05) exhibits substantial variation (min = 0.002, max = 0.38), indicating diverse approaches to expected loss absorption. The risk-weighted assets ratio (average = 0.37) highlights how banks' balance sheets are structured regarding regulatory risk exposure.

```{r esg-risk-heatmap, fig.cap="Correlation Heatmap: ESG Components vs. Risk Metrics"}
# Create a focused heatmap showing correlations between ESG components and risk metrics
esg_vars <- c("esg_score", "environmental_pillar_score", "social_pillar_score", "governance_pillar_score")
risk_vars <- c("cet1_risk_exposure", "tier1_risk_exposure", "totalcap_risk_exposure", 
               "leverage_ratio", "provisions_ratio", "rwa_ratio")

cor_matrix <- data_analysis %>%
  select(all_of(c(esg_vars, risk_vars))) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("variable1") %>%
  filter(variable1 %in% esg_vars) %>%
  select(variable1, all_of(risk_vars)) %>%
  pivot_longer(-variable1, names_to = "variable2", values_to = "correlation")

# Simplify variable names for better visualization
cor_matrix <- cor_matrix %>%
  mutate(
    variable1 = case_when(
      variable1 == "esg_score" ~ "ESG Score",
      variable1 == "environmental_pillar_score" ~ "Environmental",
      variable1 == "social_pillar_score" ~ "Social",
      variable1 == "governance_pillar_score" ~ "Governance"
    ),
    variable2 = case_when(
      variable2 == "cet1_risk_exposure" ~ "CET1 Ratio",
      variable2 == "tier1_risk_exposure" ~ "Tier 1 Ratio",
      variable2 == "totalcap_risk_exposure" ~ "Total Capital Ratio",
      variable2 == "leverage_ratio" ~ "Leverage Ratio",
      variable2 == "provisions_ratio" ~ "Provisions Ratio",
      variable2 == "rwa_ratio" ~ "RWA Ratio"
    )
  )

ggplot(cor_matrix, aes(x = variable2, y = variable1, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 3) +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#4575B4", 
                       midpoint = 0, limits = c(-0.5, 0.5)) +
  labs(x = "", y = "", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure \@ref(fig:esg-risk-heatmap) reveals distinct correlation patterns between ESG components and banking risk metrics. The overall ESG score shows negative correlations with capital adequacy measures (CET1: -0.25, Tier 1: -0.08), suggesting that stronger ESG performance might be associated with lower risk-weighted capital ratios. This counterintuitive relationship warrants deeper examination in subsequent analyses. Among ESG pillars, the social dimension demonstrates the strongest negative association with capital ratios (CET1: -0.35), while environmental and governance pillars show weaker relationships (CET1: -0.11 and -0.12 respectively). This pattern suggests heterogeneity in how different sustainability dimensions relate to risk profiles, with social performance potentially playing a more significant role in banks' risk positioning.

# Methodology

This section presents the empirical approach used to examine the relationship between ESG performance and banking risk. This paper employs several complementary methodologies to address potential issues of unobserved heterogeneity, non-linearity, and endogeneity, providing a comprehensive analysis of how ESG factors relate to various banking risk measures.

## Fixed Effects Panel Models

The primary analytical framework employs panel data models with two-way fixed effects, incorporating non-linear specifications to capture potential curvilinear relationships. This approach controls for both time-invariant bank characteristics and common time trends, addressing concerns about unobserved heterogeneity highlighted in previous banking studies (Gehrig et al., 2023; Palmieri et al., 2023).

For the overall ESG score, the non-linear specification is:
$$Y_{i,t} = \beta_0 + \beta_1 \text{ESG}_{i,t} + \beta_2 \text{ESG}_{i,t}^2 + \beta_3 Y_{i,t-1} + \boldsymbol{\beta_4} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$
where $Y_{i,t}$ represents the risk measure for bank $i$ in year $t$ (e.g., CET1 risk exposure, Tier 1 risk exposure, leverage ratio); $\text{ESG}_{i,t}$ denotes the overall ESG score; $\text{ESG}_{i,t}^2$ is the squared term capturing potential non-linear effects; $Y_{i,t-1}$ represents a lagged dependent variable to account for persistence in risk measures; $\mathbf{X}_{i,t}$ is a vector of control variables including the natural logarithm of total assets and the equity-to-assets ratio; $\alpha_i$ are bank fixed effects capturing time-invariant bank characteristics; $\lambda_t$ are year fixed effects accounting for common time trends; and $\varepsilon_{i,t}$ is the error term.

For a more granular analysis of ESG components, the model is extended to include non-linear terms for each pillar:
$$Y_{i,t} = \beta_0 + \beta_1 \text{ENV}_{i,t} + \beta_2 \text{ENV}_{i,t}^2 + \beta_3 \text{SOC}_{i,t} + \beta_4 \text{SOC}_{i,t}^2 + \beta_5 \text{GOV}_{i,t} + \beta_6 \text{GOV}_{i,t}^2 + \beta_7 Y_{i,t-1} + \boldsymbol{\beta_8} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$
where $\text{ENV}_{i,t}$, $\text{SOC}_{i,t}$, and $\text{GOV}_{i,t}$ represent the environmental, social, and governance pillar scores, respectively, with their corresponding squared terms.

This differentiated approach follows Bannier et al. (2021) and Palmieri et al. (2023), who demonstrated that individual ESG components may influence risk metrics differently. The inclusion of quadratic terms addresses potential non-linear relationships suggested by Barth et al. (2022) and Ersoy et al. (2022), while the dynamic specification with lagged dependent variables aligns with approaches used by Danisman and Tarazi (2023) and Gehrig et al. (2023) to account for gradual adjustments in bank risk profiles in response to changes in ESG factors. The fixed effects specification helps mitigate omitted variable bias from time-invariant factors, a crucial concern noted by Carnevale and Drago (2024) in ESG-banking research.

## Quantile Regression Analysis

To examine potential heterogeneous effects of ESG factors across different levels of banking risk a quantile regression analysis is employed, following methodological precedents in banking literature such as Curcio et al. (2023) and Izcan and Bektas (2022). Unlike standard OLS or fixed effects models that focus on the conditional mean, quantile regressions estimate the relationship at different points in the conditional distribution of the dependent variable:
$$Q_{Y_{i,t}}(\tau|\text{ESG}_{i,t}, \mathbf{X}_{i,t}) = \beta_0(\tau) + \beta_1(\tau) \text{ESG}_{i,t} + \boldsymbol{\beta_2}(\tau) \mathbf{X}_{i,t}$$
where $Q_{Y_{i,t}}(\tau|\text{ESG}_{i,t}, \mathbf{X}_{i,t})$ denotes the $\tau$-th conditional quantile of the risk measure $Y_{i,t}$, and $\beta_j(\tau)$ are the coefficients that may vary across quantiles. I estimate this model for five quantiles ($\tau = 0.1, 0.25, 0.5, 0.75, 0.9$), providing insights into how ESG scores affect banks with different risk profiles. This approach is valuable for regulatory purposes, as it helps identify whether ESG factors have stronger implications for higher-risk institutions.

For pillar-specific effects, I estimate:
$$Q_{Y_{i,t}}(\tau|\text{ENV}_{i,t}, \text{SOC}_{i,t}, \text{GOV}_{i,t}, \mathbf{X}_{i,t}) = \beta_0(\tau) + \beta_1(\tau) \text{ENV}_{i,t} + \beta_2(\tau) \text{SOC}_{i,t} + \beta_3(\tau) \text{GOV}_{i,t} + \boldsymbol{\beta_4}(\tau) \mathbf{X}_{i,t}$$
This allows for examining whether specific ESG components have differential impacts across the risk distribution, a nuance that means-based approaches might obscure.

## Principal Component Analysis

To address the high dimensionality and potential multicollinearity among ESG components, I employ Principal Component Analysis (PCA). This approach is particularly valuable given the detailed ESG component data available in the Refinitiv database, which includes numerous subcategories beyond the three main pillars.

PCA transforms the original ESG variables into a set of orthogonal principal components that capture the maximum variance in the data:
$$PC_j = \sum_{k=1}^p a_{jk} Z_k$$
where $PC_j$ is the $j$-th principal component, $Z_k$ represents the standardized $k$-th ESG variable, and $a_{jk}$ are the component loadings. I retain enough components to explain approximately 80% of the total variance, with a maximum of five components to maintain interpretability.

These principal components are then used as regressors in the fixed effects model:
$$Y_{i,t} = \beta_0 + \sum_{j=1}^m \beta_j PC_{j,i,t} + \boldsymbol{\beta_{m+1}} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$
where $m$ is the number of retained principal components. This PCA approach follows methodological insights from Goss and Roberts (2010), who applied similar techniques to reduce dimensionality in KLD social responsibility data. It allows for capturing the underlying structure of ESG performance while avoiding collinearity issues that might arise when including multiple correlated ESG components simultaneously.

# Results

This section presents the main results from the empirical analyses. It first present the findings from the fixed effects panel models, examining the impact of overall ESG scores and individual ESG pillars on risk-weighted capital ratios, leverage ratios, provisions, and liquidity ratios. I then discuss the results from the quantile regression analysis, highlighting how ESG factors affect banks with different risk profiles. Finally, I present the outcomes of the PCA analysis, illustrating how principal components of ESG performance relate to banking risk metrics.

## Fixed Effects Panel Models

### Overall ESG score

Table \@ref(tab:esg-table) presents the fixed effects estimates for the relationship between the overall ESG score and eight different risk measures. The models include both linear and quadratic ESG terms to capture potential non-linearities, along with control variables (log assets and equity-to-assets ratio).

```{r esg-table}
ms_esg_table
```

The results reveal a nuanced relationship between ESG performance and bank risk measures. For the capital adequacy ratios (CET1, Tier 1, and total capital risk exposure), the coefficients on the linear ESG score term are positive but statistically insignificant. The quadratic terms are similarly insignificant across these specifications, indicating no detectable non-linear relationship within the sample period.

For leverage ratio, provisions, and provisions ratio, again statistically insignificant coefficients on both the linear and quadratic ESG terms are observable. The lack of significance persists across liquidity ratio and risk-weighted assets (RWA) ratio models, though the coefficient for the squared ESG term in the RWA ratio model approaches marginal significance (p < 0.1). 

In contrast, the control variables exhibit strong and significant relationships with risk measures. Bank size (measured by log assets) positively correlates with all three capital adequacy ratios, with significant coefficients ranging from 0.185 to 0.238 (p < 0.001). This aligns with the expectation that larger banks maintain higher capital buffers, potentially due to heightened regulatory scrutiny or superior access to capital markets. Similarly, the equity-to-assets ratio demonstrates strong positive associations with capital adequacy measures, leverage ratio, and provisions ratio, with coefficients ranging from 1.327 to 4.727 (p < 0.01 or better). This reflects the mechanical relationship between equity holdings and risk-based capital ratios, as well as the broader tendency of well-capitalized banks to maintain stronger risk buffers.

These results differ somewhat from the literature, who mostly documented more consistent relationships between ESG scores and bank stability measures. However, the use of regulatory risk metrics rather than market-based measures may account for some of these differences. Additionally, the focus on a post-2019 period captures a regulatory environment with more developed ESG integration, potentially altering the ESG-risk relationship compared to earlier studies.

### ESG Pillar Score Effects

To further investigate the potential heterogeneity in how different ESG dimensions affect banking risk, Table \@ref(tab:pillar-table) presents fixed effects estimates that decompose the overall ESG score into its three constituent pillars: Environmental, Social, and Governance. 

```{r pillar-table}
ms_pillar_table
```

The results reveal distinct patterns across the three pillars. The Environmental pillar demonstrates a negative relationship with CET1 risk exposure (-0.006), though not statistically significant. However, for provisions ratio, the Environmental score shows a significant positive association (0.007, p < 0.05), with a corresponding significant negative quadratic term (-0.000, p < 0.05). This suggests that while higher environmental performance initially correlates with higher provisions, this relationship diminishes at very high levels of environmental performance—potentially indicating that moderate environmental improvements correspond to more conservative provisioning practices, but extreme environmental focus may come at the expense of such prudence.

The Social pillar exhibits the strongest associations with risk measures among the three ESG dimensions. The coefficients for Social score are positive and statistically significant for all three capital adequacy measures: 0.027 for CET1 risk exposure, 0.030 for Tier 1 risk exposure, and 0.030 for total capital risk exposure (all p < 0.01). These are accompanied by significant negative quadratic terms (-0.000, p < 0.01), indicating an inverted U-shaped relationship where the positive impact on capital ratios diminishes at very high Social score levels. This pattern aligns with findings from Ashwin Kumar et al. (2016), who documented relationships between social performance and financial stability. For RWA ratio, the Social pillar shows a significant negative linear coefficient (-0.014, p < 0.05) with a positive quadratic term (0.000, p < 0.01), suggesting a U-shaped relationship where moderate social performance correlates with lower risk-weighted assets, but this effect reverses at extreme values.

The Governance pillar demonstrates marginally significant negative associations with Tier 1 and total capital risk exposures (-0.003 and -0.004, respectively, p < 0.1), with corresponding positive quadratic terms (both 0.000, p < 0.1). This suggests a subtle U-shaped relationship where moderate governance scores correlate with lower capital ratios, but very high governance scores may actually correspond to higher capital adequacy. This non-linear pattern aligns with findings from Barth et al. (2022), who observed complex relationships between governance factors and credit risk.

Control variables maintain similar patterns to those observed in the overall ESG score models, with log assets and equity-to-assets ratio showing strong positive associations with capital adequacy measures. The model fit improves noticeably compared to the overall ESG score specifications, with R-squared values increasing across all risk measures. For instance, the R-squared for CET1 risk exposure increases from 0.457 to 0.544, suggesting that distinguishing between ESG pillars captures meaningful variation that aggregate scores obscure.

These findings highlight the importance of disaggregating ESG scores into their component pillars, as the three dimensions exhibit distinct—and sometimes opposing—relationships with banking risk measures. The consistently significant effects observed for the Social pillar suggest that social performance may be particularly relevant for bank risk profiles, potentially through mechanisms related to stakeholder trust, reputation, and employee relations as highlighted by Amiraslani et al. (2017) and Godfrey (2005).

### Limitations of the Fixed Effects Approach

However, several limitations warrant consideration. First, fixed effects models cannot account for potential reverse causality between ESG performance and banking risk. Banks with stronger capital positions might allocate more resources to ESG initiatives, creating a bidirectional relationship that fixed effects alone cannot disentangle. This concern is particularly relevant given findings by Cornett et al. (2016) and Bouslah et al. (2013) suggesting that financial performance may drive subsequent ESG investments.

Second, fixed effects estimates identify relationships based on within-bank variation over time, which may be limited in the relatively short sample period (2020-2023). The modest time variation in ESG scores—as evident in Figure \@ref(fig:time-trends)—could contribute to the limited statistical significance observed for some relationships. This temporal constraint also means the analysis primarily captures short-term associations rather than long-term equilibrium relationships between ESG factors and banking risk.

## Quantile Regression Analysis

This section presents the results from quantile regression analysis, which examines the potentially heterogeneous effects of ESG factors across different levels of banking risk. While the fixed effects models presented earlier provide insights into average relationships between ESG performance and risk metrics, they may mask important variations across the risk distribution. Quantile regression addresses this limitation by estimating the relationship at different points in the conditional distribution of the dependent variable, providing a more nuanced understanding of how ESG factors relate to banking risk profiles.

### Overall ESG Score 

Table \@ref(fig:quantile-ESG) presents the estimated effects of overall ESG scores on banking risk measures across different quantiles. The results reveal patterns that would be obscured in a conventional mean-based analysis. For CET1 risk exposure, the effect of ESG scores is consistently negative across all quantiles, ranging from -0.004 at the 0.1 quantile to -0.002 at the 0.9 quantile. This suggests that higher ESG performance is associated with lower CET1 risk exposure across the entire distribution, with a slightly stronger effect for banks with already lower capital ratios (lower quantiles).

```{r}
# Run the quantile regression analysis
results <- run_simplified_esg_analysis(
  data = data_analysis,  
  dep_vars = c("cet1_risk_exposure", "tier1_risk_exposure", 
               "leverage_ratio", "provisions_ratio"),
  tau_values = c(0.1, 0.25, 0.5, 0.75, 0.9)
)
```

```{r}
summary_data <- get_raw_summary_tables(
  results$coefficients$esg, 
  results$coefficients$pillars
)
```

```{r quantile-esg-table}
make_simple_esg_summary_table(summary_data)
```

The statistical significance of these effects varies across the distribution. As shown in Table \@ref(tab:quantile-esg-table), the ESG coefficient is significant at the 5% level for the 0.1 quantile (-0.004, p < 0.05) and at the 0.1% level for the 0.25 and 0.5 quantiles (-0.003, p < 0.001). However, it becomes only marginally significant at the 0.75 quantile (-0.003, p < 0.1) and insignificant at the 0.9 quantile (-0.002, p > 0.1). This pattern suggests that ESG performance has a more reliable relationship with capital adequacy for banks in the lower to middle ranges of the risk distribution.

For Tier 1 risk exposure, a similar pattern emerges, with a negative relationship that is most pronounced at the median quantile. The effect becomes more varied at the extremes of the distribution, particularly at the 0.9 quantile where the confidence interval widens substantially. This indicates greater uncertainty in the ESG-risk relationship for banks with the highest Tier 1 risk exposure ratios.

The relationship between ESG scores and leverage ratio shows minimal variation across quantiles, with coefficients close to zero throughout the distribution. This aligns with the findings from fixed effects models, suggesting that ESG performance has limited direct association with leverage decisions.

For provisions ratio, the coefficients are slightly positive and relatively stable across quantiles, though generally lacking statistical significance. This suggests that while higher ESG scores might be associated with marginally higher provisioning, this relationship is not particularly strong at any point in the risk distribution.

```{r cet1-quantile-esg}
make_esg_table(results$coefficients$esg, "cet1_risk_exposure")
```

Table \@ref(tab:cet1-quantile-esg) provides a detailed view of quantile regression results for CET1 risk exposure across different quantiles. Beyond the ESG coefficient, it reveals interesting patterns in control variables. The effect of log assets on CET1 risk exposure decreases monotonically across quantiles, from 0.027 (p < 0.05) at the 0.1 quantile to -0.040 (p < 0.05) at the 0.9 quantile. This suggests that bank size has a positive relationship with capital adequacy for riskier banks (lower quantiles) but a negative relationship for safer banks (higher quantiles), a nuanced finding that would be missed in mean-based analyses.

Similarly, the coefficient for equity-to-assets ratio shows substantial variation, from a positive but insignificant 1.096 at the 0.1 quantile to a negative and marginally significant -2.294 (p < 0.1) at the 0.9 quantile. This unexpected pattern at higher quantiles might reflect that banks with already high CET1 ratios exhibit different strategic approaches to overall equity levels, potentially optimizing other aspects of their capital structure.

### ESG Pillar Effects Across Risk Quantiles

Disaggregating ESG into its component pillars reveals even more nuanced relationships that vary across the risk distribution, as shown in Figures \@ref(fig:quantile-environmental), \@ref(fig:qunatile-social), and \@ref(fig:quantile-governance) in the Appendix and table \@ref(tab:quantile-pillar-table). 

```{r quantile-pillar-table}
make_simple_pillar_summary_table(summary_data)
```

The environmental pillar demonstrates a relationship with CET1 risk exposure that varies substantially across quantiles. At lower quantiles (0.1 and 0.25), the relationship is weakly negative, becoming more positive at higher quantiles, though generally not reaching statistical significance (Table \@ref(tab:quantile-pillar-table)). This pattern suggests that environmental performance may have different implications for banks depending on their position in the risk distribution, with potentially beneficial effects for lower-risk banks.

For provisions ratio, the environmental score shows an interesting pattern where the relationship is initially positive at lower quantiles but becomes increasingly negative at higher quantiles. This could indicate that banks with already higher provisioning levels (upper quantiles) that focus on environmental performance might adopt more conservative provisioning practices, while the opposite occurs for banks with lower provisions.

The social pillar exhibits the most consistent relationship with capital adequacy measures across the risk distribution. For CET1 risk exposure, the coefficient is negative across all quantiles, with values ranging from -0.004 at the 0.25 quantile to -0.008 at the 0.75 quantile (Table \@ref(tab:quantile-pillar-table)). The relationship is statistically significant at the 0.1% level for both the 0.25 and 0.5 quantiles, becoming even stronger at the 0.75 quantile. This pattern suggests that social performance has a particularly strong negative association with capital risk for banks in the middle to upper-middle portions of the risk distribution.

This finding aligns with and extends the results from Ashwin Kumar et al. (2016), who documented relationships between social performance and financial stability. While they found a general association, the quantile analysis reveals that this relationship is more pronounced for banks with moderate to higher capital buffers, suggesting a potential complementarity between social responsibility and prudent capital management for these institutions.

The governance pillar demonstrates the most varied relationship across the risk distribution. For CET1 risk exposure, the coefficient is close to zero at most quantiles, but shows a modest positive relationship at the 0.75 quantile (Table \@ref(tab:quantile-pillar-table)). For provisions ratio, the relationship is slightly positive at median quantiles but becomes negative at extreme quantiles. This heterogeneity aligns with findings from Barth et al. (2022), who observed complex relationships between governance factors and risk metrics.

```{r leverage-quantile-esg}
make_pillar_table(results$coefficients$pillars, "leverage_ratio")
```

Table \@ref(tab:leverage-quantile-esg) presents detailed results for the pillar-specific analysis of leverage ratio. Notably, the environmental score shows a statistically significant negative relationship at the 0.25 quantile (-0.001, p < 0.01), but this effect disappears at other points in the distribution. The social and governance pillars show minimal effects across all quantiles. Control variables, particularly equity-to-assets ratio, exhibit strong positive associations with leverage ratio across all quantiles, with coefficients increasing monotonically from 2.193 (p < 0.001) at the 0.1 quantile to 3.010 (p < 0.001) at the 0.9 quantile. This pattern reflects the mechanical relationship between equity holdings and regulatory capital requirements, with the effect becoming stronger for banks with already higher leverage ratios.

## Principal Component Analysis

This section presents results from Principal Component Analysis (PCA) of ESG dimensions and their relationship with banking risk. The first five components explain approximately 83% of total variance in ESG variables, with the first component alone accounting for 46.2% followed by the second with 13.8%. This substantial dimensionality reduction suggests that while ESG performance is multidimensional, relatively few underlying factors capture most variation across European banks.

```{r}
pca_results <- run_pca_esg_analysis(
  data = data_analysis,  
  dep_vars = c("cet1_risk_exposure", "tier1_risk_exposure", 
               "totalcap_risk_exposure", "leverage_ratio", "provisions", 
               "provisions_ratio", "liquidity_ratio", "rwa_ratio"),
  variance_threshold = 0.8
)
```

Figure \@ref(fig:pca-loading-plot) reveals the composition of these principal components. PC1 shows negative associations with multiple ESG variables (overall ESG score, ESG score squared, social score, social score squared, and governance score squared), representing a general ESG factor capturing common variance across sustainability dimensions. Essentially, PC1 functions as an inverse measure of general ESG quality.

```{r pca-loading-plot}
pca_results$visualizations$loading_plot
```

PC2 demonstrates a more nuanced pattern, positively associating with "ESG Combined" and "Product Responsibility" while negatively associating with environmental measures. This suggests PC2 contrasts social and product responsibility against environmental considerations, potentially capturing trade-offs between these dimensions in banks' sustainability strategies.

As shown in Table \@ref(tab:pca-interpretation), PC3 (10.2% of variance) juxtaposes product responsibility and social performance against management quality and governance, while PC4 (7.2%) captures ESG controversies and innovations versus product responsibility. PC5 (5.4%) contrasts emissions and product responsibility against human rights and community engagement.

```{r pca-interpretation}
pca_results$visualizations$pc_interpretation
```

These component structures extend previous research on ESG performance's multidimensional nature. The component specifically related to controversies (PC4) is particularly noteworthy given recent literature highlighting ESG controversies' importance in financial stability (Amiraslani et al., 2017; Galletta et al., 2022).

Table \@ref(tab:pca-table) presents fixed effects estimates for relationships between extracted principal components and banking risk measures.

```{r pca-table}
pca_results$visualizations$pc_model_table
```

PC1 (inverse ESG quality) shows weak relationships with capital measures, though none reach statistical significance. PC2 (social/product responsibility vs. environmental focus) demonstrates consistently negative relationships with capital adequacy measures, suggesting banks emphasizing social aspects over environmental ones may maintain somewhat lower capital ratios.

PC3 (stakeholder engagement vs. governance) exhibits the strongest relationships with capital adequacy measures (-0.010 for CET1, -0.007 for Tier 1, -0.009 for total capital risk exposure). While not statistically significant, these consistent relationships suggest banks with stronger external stakeholder engagement relative to governance might maintain lower capital ratios, aligning with Di Tommaso and Thornton's (2020) findings.

For provisions ratio, PC2 shows a marginally significant positive relationship (0.010, p < 0.1), indicating banks emphasizing social responsibility maintain higher provisions, potentially offsetting lower capital ratios. PC4 and PC5 show varied relationships without statistical significance, possibly reflecting their specialized nature.

Control variables maintain patterns similar to previous analyses, with bank size positively correlating with capital adequacy measures (coefficients 0.199-0.261, p < 0.001) and equity-to-assets ratio showing strong positive associations across risk measures (coefficients 1.771-4.611, p < 0.001).

These PCA results offer key insights beyond aggregate ESG analyses. First, they confirm ESG performance's multidimensionality, with distinct underlying factors exhibiting different relationships with risk measures. Second, components cutting across traditional ESG categories—such as PC2 contrasting social responsibility against environmental considerations—suggest potential trade-offs in banks' sustainability approaches with different implications for risk management and capital allocation.

# Conclusion

This study provides empirical evidence on the relationship between ESG performance and banking risk for European banks from 2020 to 2023, employing multiple methodological approaches to address potential concerns of unobserved heterogeneity, non-linearity, and heterogeneous effects across the risk distribution. The results reveal several important insights for researchers, practitioners, and policymakers interested in the intersection of sustainability and financial stability.

First, while the overall ESG score demonstrates limited association with risk metrics in fixed effects specifications, the disaggregation into component pillars uncovers substantial heterogeneity in their effects. The social pillar consistently emerges as the most influential ESG dimension, exhibiting significant positive relationships with capital adequacy measures characterized by inverted U-shaped patterns. This finding suggests that banks' engagement with social considerations—including employee relations, community involvement, and human rights—may be particularly relevant for risk management practices, potentially operating through channels of reputational capital and stakeholder trust.

Second, the quantile regression analysis reveals that ESG effects vary significantly across the risk distribution. The overall ESG score demonstrates a consistently negative relationship with CET1 risk exposure that is strongest for banks in the lower to middle quantiles, while the social pillar exhibits the most consistent negative association with capital adequacy measures across the entire distribution. These findings highlight the value of moving beyond mean-based approaches when examining the ESG-risk relationship, with significant implications for supervisory focus and regulatory design.

Third, the principal component analysis uncovers a more nuanced underlying structure of ESG performance that extends beyond the traditional environmental, social, and governance pillars. The identification of components that cut across conventional ESG categories—such as the contrast between social responsibility and environmental considerations (PC2), or the juxtaposition of stakeholder engagement against governance (PC3)—suggests that banks may face strategic choices in their sustainability approaches, with different implications for risk profiles.

Nevertheless, this study faces several limitations that warrant consideration. The relatively short time period (2020-2023) limits the ability to identify long-term equilibrium relationships between ESG factors and banking risk. Additionally, despite the methodological approach, concerns about potential reverse causality remain, as banks with stronger capital positions might allocate more resources to ESG initiatives, creating bidirectional relationships that fixed effects cannot fully disentangle.

Future research could address these limitations by employing longer time series, particularly as ESG reporting continues to mature. Researchers should consider instrumental variable approaches to more directly address endogeneity concerns, possibly leveraging regulatory shifts or external shocks as exogenous variation in ESG focus. Additionally, investigating the channels through which ESG factors—particularly social dimensions—influence banking risk represents a promising direction, potentially through customer loyalty, employee retention, or regulatory anticipation. Finally, as ESG disclosure requirements continue to evolve under initiatives like the EU Taxonomy and the Corporate Sustainability Reporting Directive, future studies could examine how standardization in reporting affects the relationship between sustainability performance and banking stability.

In conclusion, this study contributes to the understanding of the ESG-risk nexus in banking by highlighting the multidimensional nature of sustainability performance and its varied relationships with risk metrics across different bank profiles. These findings underscore the potential value of integrating nuanced ESG considerations into both risk management practices and regulatory frameworks, while acknowledging the complex trade-offs that may exist between different sustainability dimensions.

# Appendix

## References

## Tables

```{r variables-table, echo=FALSE}
source("variables_df.R")
kable(variables_df, 
      caption = "Overview of variables", 
      booktabs = TRUE,
      longtable = TRUE) %>%
  kable_styling(latex_options = c("striped", "repeat_header", "hold_position"),
                font_size = 9) %>%
  column_spec(1, bold = TRUE, width = "2.5in") %>%
  column_spec(2, width = "2.5in") %>%
  column_spec(3, width = "1in")
```

```{r distributions, echo=FALSE, fig.cap="Distributions of Key Variables."}
# Define variables for distribution plots
dist_vars <- c("esg_score", "cet1_risk_exposure", "leverage_ratio",
               "log_assets", "equity_to_assets") # Choose a few important ones

# Generate the plot
distribution_plot <- generate_distribution_plots(data_analysis, dist_vars)

# Print the plot to include it in the document
print(distribution_plot)
```


```{r risk-size-table, echo=FALSE, results='asis', fig.cap="Risk Metrics by Bank Size Quartile"}
# Create a new table grouping banks by size quartiles and showing risk metrics
data_analysis %>%
  mutate(size_quartile = ntile(log_assets, 4)) %>%
  group_by(size_quartile) %>%
  summarize(
    `Number of Banks` = n(),
    `Avg Total Assets (€ Billion)` = mean(exp(log_assets))/1000,
    `CET1 Risk Exposure` = mean(cet1_risk_exposure, na.rm=TRUE),
    `Tier 1 Risk Exposure` = mean(tier1_risk_exposure, na.rm=TRUE),
    `Leverage Ratio` = mean(leverage_ratio, na.rm=TRUE),
    `Provisions Ratio` = mean(provisions_ratio, na.rm=TRUE),
    `RWA Ratio` = mean(rwa_ratio, na.rm=TRUE)
  ) %>%
  kable(caption = "Risk Metrics by Bank Size Quartile",
        digits = 3,
        col.names = c("Size Quartile", "N", "Avg Assets (€B)", 
                      "CET1 Ratio", "Tier 1 Ratio", "Leverage Ratio",
                      "Provisions Ratio", "RWA Ratio"),
        booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r summary-stats, echo=FALSE, results='asis', fig.cap="Descriptive Statistics"}
# Define the variables you want to summarize
key_vars_summary <- c("esg_score", "environmental_pillar_score", "social_pillar_score",
                      "governance_pillar_score", "cet1_risk_exposure", "tier1_risk_exposure",
                      "totalcap_risk_exposure", "leverage_ratio", "provisions",
                      "provisions_ratio", "liquidity_ratio", "rwa_ratio",
                      "log_assets", "equity_to_assets") # Select relevant vars

# Generate and print the table
summary_table <- generate_summary_table(data_analysis,
                                        vars = key_vars_summary,
                                        title = "Descriptive Statistics of Key Variables")
print(summary_table)
```


Table \@ref(tab:correlation-matrix) displays the Pairwise Pearson correlation patterns among key variables, revealing several notable relationships. The overall ESG score exhibits strong positive correlations with the social (0.83) and governance (0.85) pillars, and a moderate correlation with the environmental pillar (0.62). Interestingly, ESG scores show negative correlations with risk metrics: specifically, -0.25 with CET1 risk exposure. This negative relationship, though modest, aligns with the hypothesis that stronger ESG performance may be associated with lower risk levels.

The three ESG pillars show different patterns of association with risk metrics. The environmental pillar correlates negatively (-0.11) with CET1 risk exposure, while the social pillar shows a stronger negative relationship (-0.35). Governance displays the weakest correlation with CET1 risk (-0.12). Bank size (log assets) demonstrates positive correlations with ESG score (0.50) and environmental pillar score (0.58), suggesting that larger banks tend to have stronger overall ESG performance, particularly in environmental areas. This pattern might reflect larger institutions' greater resources for sustainability initiatives or their enhanced regulatory scrutiny and reporting requirements. Notably, the leverage ratio shows a negative correlation with log assets (-0.45), indicating that larger banks tend to operate with relatively lower leverage ratios.

```{r correlation-matrix, echo=FALSE, results='asis', fig.cap="Correlation Matrix."}
# Define variables for the correlation matrix (can be a subset for readability)
cor_vars <- c("esg_score", "environmental_pillar_score", "social_pillar_score",
              "governance_pillar_score", "cet1_risk_exposure", "leverage_ratio",
              "provisions_ratio", "liquidity_ratio", "rwa_ratio",
              "log_assets", "equity_to_assets")

# Generate and print the correlation table
correlation_table <- generate_correlation_table(data_analysis,
                                                vars = cor_vars,
                                                title = "Correlation Matrix")
print(correlation_table)
```

## Time Trends

Figure \@ref(fig:time-trends) depicts the evolution of four key variables from 2020 to 2023, revealing important temporal patterns. The ESG score shows an overall positive trajectory, rising from approximately 73.1 in 2020 to 73.9 in 2023, with a temporary decline to 72.5 in 2021. This pattern suggests that despite potential pandemic-related setbacks in 2021, European banks have generally strengthened their sustainability practices over the observed period.
CET1 risk exposure displays a contrasting trend, declining from 0.67 in 2020 to approximately 0.60 in 2021-2022, before increasing slightly to 0.61 in 2023. This initial decrease might reflect regulatory flexibility during the pandemic or strategic shifts in banks' risk-weighted asset compositions. The subsequent stabilization and slight uptick suggest a potential return to pre-pandemic risk levels.
The leverage ratio exhibits more volatility, increasing from 0.21 in 2020 to 0.24 in 2021, declining to 0.23 in 2022, then rising again to 0.24 in 2023. This fluctuation may reflect banks' adaptive responses to changing economic conditions and regulatory expectations throughout the post-pandemic recovery period.
Liquidity ratio shows the most dramatic shift, rising sharply from 0.12 in 2020 to approximately 0.16 in 2021-2022, before declining to 0.14 in 2023. The initial increase likely represents a prudential response to pandemic-induced uncertainty, with banks building liquidity buffers. The subsequent decline might indicate a gradual return to more normalized liquidity management as economic conditions stabilized.

Collectively, these temporal trends reveal a complex adaptive landscape where banks have generally improved their ESG performance while navigating significant macroeconomic challenges that affected their risk and capital metrics. The non-uniform patterns across different variables highlight the multifaceted nature of banks' strategic responses to changing regulatory, economic, and sustainability pressures during this period.

```{r time-trends, echo=FALSE, fig.cap="Average Trends Over Time."}
# Define variables for time trend plots
trend_vars <- c("esg_score", "cet1_risk_exposure", "leverage_ratio", "liquidity_ratio")

# Generate the plot
time_trend_plot <- generate_time_trend_plots(data_analysis, trend_vars)

# Print the plot
print(time_trend_plot)
```

## Quantile regression

```{r quantile-ESG}
results$plots$combined$esg
```

```{r quantile-environmental}
results$plots$combined$environmental
```

```{r qunatile-social}
results$plots$combined$social
```

```{r quantile-governance}
results$plots$combined$governance
```

## PCA

```{r pca-results}
pca_results$visualizations$scree_plot
```