---
title: "ESG and banking risk"
author: "Maurits de Haan (s8161658, m.c.p.dehaan@students.uu.nl)"
date: | 
  | 3rd of April, 2025
  | Course: Data Science Lab (ECB3DSL), Academic year 2024/2025
  | Individual internship for FNO Zorg voor Kansen
  | Supervisor: Dr. Ulrich Zierahn-Weilage
  | Word Count: XXXX (max 5000)
output:
  bookdown::pdf_document2:
      toc: false
      toc_depth: 3
      number_sections: true
      fig_height: 4
      fig_width: 6
      includes:
        in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)
```

```{r}
library(knitr)
library(kableExtra)
```

```{r}
# Load required libraries
library(plm)
library(dplyr)
library(stargazer)
library(kableExtra)
library(modelsummary)
library(broom)
library(ggplot2)
library(gridExtra)
library(plm)
library(AER)
library(gridExtra)
library(kableExtra)
library(modelsummary)
library(plm)
library(dplyr)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(patchwork)
library(modelsummary) # Already used in your example
library(broom)        # Already used in your example


sink(file = tempfile())
invisible(source(here::here("R", "import_ESG.R")))
invisible(source(here::here("R", "import_financial.R")))
invisible(source(here::here("R", "merging_ESG_financial.R")))
invisible(source(here::here("R", "analysis_fixed_effects_quadratic.R")))
invisible(source(here::here("R", "analysis_quantile.R")))
invisible(source(here::here("R", "analysis_pca.R")))
invisible(source(here::here("R", "analysis_iv.R")))
invisible(source(here::here("R", "analysis_exp.R")))
sink()
```

# Abstract {.unnumbered}
[write here]

| Statements |
|------------------------------|
| **Statement of Originality** | 
| I, the aforementioned student, herewith declare to have written this document and that I am responsible for the content of it. I declare that the text and the work presented in this document is original and that no sources other than those mentioned in the text and its references have been used in creating it. |
| Utrecht University School of Economics is responsible solely for the supervision of completion of the work, not for the content. |
| All questions about this paper can be directed to the author. | 
|------------------------------|
| **Disclosure Statement** |
| In this project, I have made use of the following Generative AI tools: ChatGPT 4o (reasoning), Claude 3.5 Sonnet (coding) |
| ChatGPT 4o was used to help with reasoning about the methodology that would be usefull to use and the way in which this can be implemented. It was used as an extension of my own knowledge and as an external check on the assumptions I made about certain methodologies. Claude 3.5 Sonnet was utilised for certain coding tasks, mostly when it was outside of the scope of what I have learned earlier. All code that was provided by the model was thouroughly checked and validated if it was actully true (and in most cases I had to rewrite it completely). Next to that, Claude was used to integrate some functions in the code that I wrote. Claude was also used to help with for example a .tex file to include the image in the header of this file and other "nice-to-have" things. |
| All content in this report was written by myself and not by ChatGPT. Most was first written by hand and then typed in RStudio. | 
|------------------------------|
| **Signature**
| |


\newpage 

\tableofcontents

\newpage

# Introduction
In recent years, the role of environmental, social, and governance (ESG)  considerations in banking has gained significant attention in both academic and policy fields. Growing concerns about climate change, inequality and good governance have increased the interest how corporations and banks through their lending decision can drive the transition to a more sustainable and equitable economy (Carnevale & Drago, 2024; European Banking Authority, 2018; EU, 2014). Many banks across Europe are subject to Pillar 3 disclosures under the Capital Requirements Regulation (CRR), and new ESG disclosures further highlight climate and sustainability exposures (De Barros Fritz & Gentile, 2024). Policymakers and financial experts argue that improved ESG performance may help banks reduce their overall risk exposure, particularly by enhancing their reputation, improving stakeholder trust, and anticipating regulatory changes. However, whether ESG practices truly mitigate banking risk or whether lower-risk institutions merely find it easier to allocate resources toward socially responsible projects remains an open empirical question. This paper contributes to this debate by focusing on capital adequacy which reflects banking risk and using regulatory data instead of annual reports. In this paper, the relationship between ESG performance and capital ratios for European banks from 2020 to 2023 is analysed.

From a societal perspective, understanding whether strong ESG engagement leads to more resilient financial institutions carries high relevance. If banks with robust ESG strategies exhibit lower risk, this supports integrating ESG considerations into prudential regulation and highlights banks’ capacity to contribute to broader societal objectives without jeopardizing their stability. Conversely, if safer banks are simply better able to “afford” ESG investments, this suggests a different policy focus (e.g. financial health determines ESG commitments needed to be made). From a scientific standpoint, this study adds nuance to a growing literature that often uses conventional balance-sheet data to proxy risk, which may not fully capture regulatory metrics. By leveraging publicly available data from the European Banking Authority (EBA) , we gain a more reliable view of risk-weighted assets, Tier 1 capital, leverage, and provision levels. Pairing these risk metrics with granular ESG data from LSEG Refinitiv further strengthens the empirical inquiry, because the ESG scores (and their component pillars) benefit from refined post-2019 methodologies.

Based on the financial and ESG dataset, this paper employs a panel-data approach that exploits the ‘short-longitudinal’ structure of our dataset, comprising approximately 100 unique Eurozone banks, covering over 400 bank-year observations between 2020 and 2023. Fixed effects models allow us to control for unobserved heterogeneity at both the bank and time level, reducing omitted-variable bias. We further investigate potential non-linear effects by adding squared ESG terms, conduct lagged analyses to explore dynamic responses between ESG changes and future banking risk, and test for heterogeneous effects with quantile regressions. Finally, a principal component analysis on ESG sub-scores (see Figure X) provides an alternative way to capture the underlying dimensions of ESG and investigate their separate impacts on various risk metrics.

A brief overview of my main findings indicates the following patterns. First, in a baseline fixed-effects analysis pooling all banks, the overall ESG score shows only a modest and sometimes statistically insignificant relationship with CET1, Tier 1, and total capital ratios when using conventional robust standard errors. However, once we employ Driscoll-Kraay corrections (which account for cross-sectional dependence), a positive and significant effect emerges between ESG performance and CET1 ratios. Furthermore, restricting attention to banks with complete data over the four-year period yields stronger evidence of a beneficial ESG effect on capital adequacy. Decomposing ESG into its environmental, social, and governance pillars suggests that governance, and in some cases social-related strategies (such as corporate social responsibility), may drive much of the observed relationship—albeit with varying signs and significance across different specifications. Extended analyses (including quantile regressions and principal component analysis) reveal that individual components, particularly governance- and CSR-related dimensions, tend to be most strongly associated with capital adequacy measures.

[conclusion of Intro]

# Literature Review

The growing incorporation of environmental, social, and governance (ESG) considerations in banking decision-making has led to a growing body of research analysing whether and how ESG practices shape financial risk and performance. 

## ESG and firm capital

Fundamental papers often focus on corporate bond markets, demonstrating that firms with stronger ESG profiles benefit from lower cost of debt and better credit ratings. For instance, Apergis et al. (2022) report that lower ESG ratings tend to increase the default risk premium charged by investors, thus suggesting that ESG factors influence the pricing of unsecured corporate debt.

Further connections between ESG metrics and cost of capital have been explored by Chava (2014), who finds that firms who are involved in many environmental controvocies are penalized with higher equity and debt costs. While the magnitude of this penalty varies across subcomponents of ESG, it does suggest that, overall, markets incorporate sustainability signals into the required returns on capital. Yet, other scholars remain more cautious about conclusive results. Menz (2010), studying European corporate bonds, finds no strong link between corporate social responsibility (CSR) and secondary market credit spreads, an outcome possibly attributed to the incomplete integration of ESG in bond pricing at the time. Similarly, Amiraslani et al. (2017) observe that the positive effect of strong social capital emerges primarily in times of crisis, such as during the 2008–2009 financial turmoil, when trust and default risk dominate market concerns. They show that high-social-capital firms could secure funding at lower costs during the height of the crisis but found no pronounced link between social capital and bond spreads in more stable years.

In parallel with bond-market research, work on the relationship between ESG factors and bank loan pricing has grown. Goss and Roberts (2010) detail how banks acquire additional information from corporate ESG ratings, which translates to higher loan spreads on firms with CSR concerns. Banks effectively penalize borrowers with shortcoming ESG profiles unless those loans are well secured (Goss & Roberts, 2010). Subsequent reviews corroborate that higher ESG risk can lead to elevated bank lending costs (Carnevale & Drago, 2024). However, while evidence of this ESG “penalty” is common, differences often arise across ESG pillars, with governance sometimes playing a more substantial role in lowering perceived borrower risk.

## ESG and banking lending 

A parallel literature narrows its scope to banks themselves, examining whether ESG performance among banking institutions correlates with risk proxies such as default risk, nonperforming loans, tail risk, and market-based measures of capital adequacy. Curcio et al. (2024), for instance, use a sample of European banks and insurers to show that stronger ESG performance in nonfinancial counterparties reduces banks’ overall systemic risk exposure, albeit with nuanced differences when environmental and social facets are considered separately. This finding aligns with broader arguments that strong ESG engagement can enhance reputational capital, trustworthiness, and stakeholder relationships, all of which mitigate operational and credit risk [SOURCE].

Empirical efforts to measure direct effects of ESG on bank-specific outcomes often rely on risk-weighted capital ratios or default probability metrics. Several papers employ Merton-style distance to default or Z-scores to capture how ESG performance might reduce a bank’s vulnerability to extreme negative shocks (Chiaramonte et al., 2022; Danisman & Tarazi, 2023). The emerging consensus suggests that ESG considerations relate meaningfully to bank resilience. However, it remains unclear how the pathways are caused (e.g. via lower loan losses, improved reputational capital, or heightened stakeholder trust) (Gehrig et al., 2023). There is also extensive research focusing on operational risk: Galletta et al. (2022) document that higher ESG scores consistently reduce banks’ operational loss events, reflecting lower reputational and compliance risk. Indeed, ESG issues can directly affect operational strategies, especially during evolving regulatory standards and public scrutiny of sustainability issues.

Despite a shared recognition that sustainability dimensions matter for financial risk of banks, important theoretical and empirical debates persist if this relationship can be identified as causal. One strand, often related to stakeholder theory, proposes that deliberate ESG initiatives improve a bank’s reputation, stakeholder satisfaction, and regulatory preparedness (Godfrey, 2005; He et al., 2021). Consequently, these banks have stronger buffers against unforeseen market stresses, which may translate into stronger capital ratios or lower cost of funding. Another line of inquiry, however, posits that less risky institutions simply have the resources and latitude to engage in more robust ESG activities (Bouslah et al., 2013; Cornett et al., 2016). In that scenario, risk is arguably a precursor, rather than an outcome, of sustainability engagement. Studies such as Gehrig et al. (2023) and Danisman and Tarazi (2023) attempt to mitigate such endogeneity and simultaneity concerns by using instrumental variables or dynamic panel methodologies. 

## ESG Pillar scores

Researchers have further highlighted that different ESG pillars do not necessarily exert uniform effects on a bank’s capital adequacy. The governance dimension, for instance, has repeatedly been identified as a strong predictor of stability because it reflects managerial skill, board structure, and the capacity to address emerging sustainability concerns systematically (Chodnicka-Jaworska, 2021; Cohen, 2023). Environmental scores may be particularly relevant for banks exposed to climate-sensitive sectors, while the social dimension exhibits more mixed results [SOURCE]. For example, some studies show minimal or inconsistent links between social scores and measures such as credit spreads or default risks (Amiraslani et al., 2017; Izcan & Bektas, 2022), whereas others find that employee relations and community engagements can bolster goodwill and mitigate downside risk in stress periods (Ashwin Kumar et al., 2016).

Overall, the literature indicates that ESG has become a critical component in assessing and shaping banks’ risk profiles. Heightened stakeholder, regulatory, and market pressures ensure that ESG considerations pervade more areas of banking strategy, including the allocation of capital and management of risk-weighted assets. While a growing number of studies point to ESG factors lowering the likelihood of severe capital depletion, conclusive evidence about directionality and pillar-specificity remains somewhat inconclusive. 

# Data 

The dataset employed in this study combines two sources of bank-level information for EBA member countries: regulatory bank data and ESG data. This paper focusses on European banks that are a member of the regulatory institution the European Banking Authority (EBA). The member countries include all Eurozone counties, apart from Croatia and Slovakia, non-Eurozone countries that are transitioning to adopting the euro (Bulgaria, Hungary, Poland, Romania and Sweden), Denmark (opt-out for Eurozone participation). Next to that, the EBA also has some non-EU countries as members, such as Norway, Iceland and Liechtenstein.  These countries are all included in the initial data-gathering. For all countries, the included banks are those with their headquarter in that country. 

## Regulatory bank data

For the regulatory financial data, the EBA data from the EU-Wide Transparency Exercise is used. This is validated date by banks and EU supervisors (official regarded to as "Historical KM"). The total EBA dataset include 100 banks, with a total of 400 bank-year observations. The data is available from the third quarter of 2019 onwards to the third quarter of 2024. Since the ESG data is in full years, only the data from 2020 to 2023 is used and the yearly values are calculated. This data contains standardized regulatory metrics for Eurozone banks, including, among others, total assets, total equity, loan and advance amounts, provisions, and various capital ratios. Originally the data contained over 12,000 observations, since each bank's financial metric was recorded as a separate observation. The data was therefore pivoted wider, so that a more structures and easier to merge dataset originates. Each financial variable now has its own column. Next to that, several financial ratios were calculated, for example the equity to assets, loan to assets and provision ratio. All variables in the dataset can be found in \@ref(tab:variables-table) in the appendix. 

## ESG data

The Refinitiv database of the London Stock Exchange Group (LSEG) was used for collecting the ESG ratings. Refinitiv is the most widely used database in academic and operational analysis about sustainability ratings. ESG ratings and reporting practices have undergone refinement in recent years, especially after 2019 (Mandas et al., 2024; Fitch, 2022; Moody’s, 2022; S&P Global, 2022), implying that more recent data are likely to capture richer and more standardized information about sustainability. Papers examining European banks from 2020 onward can thus rely on these improved ESG data sets (Curcio et al., 2024; Sun et al., 2024). Such refinements are crucial because risk-related metrics, especially regarding capital buffers, hinge on regulatory definitions and supervisory guidelines, which have also evolved to incorporate ESG-related climate stress testing (BCBS, 2020; EBA, 2021).

During the data retrieving process, first all public and private companies were filtered on ICB sector code (Banking, 301010) and then on the countries where these banks have their headquarters corresponding to the EBA financial dataset. Then, the ESG variables were selected. This was done broadly, so the general ESG scores, pillar scores and component scores were all included. This namely makes it possible to execute principal component analysis later. The ESG scores were retrieved as scores (range 1-100, higher is better) rather than letter-grade ratings. This was done to make quantitative analysis more precise and understandable. Also, the NULL variables were changes to NA, the year columns were changes to rows (so that each bank in a certain year has their own row) and variable names were uniformed. 

## Merging 

The merged dataset was constructed by performing an inner join between the regulatory data (EBA) and ESG data (Refinitiv). This merging approach ensured that only banks and years appearing in both datasets are retained for the final analytical sample. This ensures that each record in the final dataset contains both the regulatory capital and risk information and the ESG performance indicators for a given institution-year observation. The two datasets were merged on the bank name and the year. Since the bank names differ between the two datasets, the Legal Entity Identifier (LEI) was used to merge the datasets. 

# Descriptive Analyses

This section provides an overview of the data used in the analysis, covering the period 2020-2023 for approximately 100 European banks in the EBA dataset. It presents descriptive statistics for the main variables of interest, examines their correlations, and illustrates their distributions and trends over time.

## Summary Statistics

Table \@ref(tab:summary-stats) presents descriptive statistics for the primary variables of interest. Our sample consists of 132 bank-year observations across the four-year period. The ESG scores exhibit considerable variation, with the overall ESG score averaging 73.32 (on a scale of 0-100, with the higher the better) and a standard deviation of 11.95 (so the majority tend to cluster around the average). Among the three ESG pillars, banks perform strongest in environmental metrics (mean = 85.98), followed by social (75.36) and governance (68.62). This pattern reflects the European banking sector's significant emphasis on environmental priorities in recent years, likely driven by EU policy initiatives such as the European Green Deal and the EU Taxonomy Regulation.

The governance pillar displays the highest variability (standard deviation = 18.94), indicating substantial heterogeneity in governance practices across European banks. This variation suggests that while environmental standards have become relatively more harmonized, governance structures and practices remain diverse across institutions.

Regarding the risk-weighted capital measures, the dataset shows that banks hold, on average, a Common Equity Tier 1 (CET1) risk exposure ratio of 0.60, with Tier 1 and total capital risk exposures standing at 0.67 and 0.76, respectively. These ratios capture how much high-quality capital (e.g., common stock and retained earnings, plus additional Tier 1 or total regulatory capital) a bank maintains relative to its risk-weighted assets. By exceeding regulatory minima, most institutions in our sample signal relatively prudent capital management practices, although the minimum values (ranging from 0.29 to 0.35) reveal that certain banks operate with tighter buffers, potentially reflecting more aggressive asset compositions or narrower margins for absorbing losses.

Turning to the leverage ratio—an unweighted measure of capital adequacy that compares Tier 1 capital against total on- and off-balance-sheet exposures—it averages 0.24 in our sample, with modest dispersion (standard deviation = 0.06). This indicates that while most banks maintain similar levels of Tier 1 capital relative to overall exposure, a minority of institutions deviate significantly in either more or less conservative directions.

Provisions, which represent funds set aside to cover expected credit losses, display substantial variation (standard deviation = 10,001.15), reflecting differences in bank size, credit risk appetite, and loan portfolio composition. When normalized as a provisions ratio (average = 0.05), these differences remain pronounced (minimum = 0.002, maximum = 0.38), signaling diverse approaches to loan-loss absorption across institutions. The liquidity ratio (mean = 0.15) serves as an indicator of the share of liquid assets available to meet short-term obligations, whereas the risk-weighted assets (RWA) ratio (average = 0.37) measures how much of a bank’s balance sheet is exposed to assets that require regulatory capital based on their riskiness, thus highlighting each bank’s overall risk management strategy.

Bank size, as captured by the natural log of total assets (mean = 13.49), underscores the mix of medium-sized and large European banks in our sample, with values ranging from 10.71 to 16.16. Finally, the equity-to-assets ratio averages 0.07, reflecting the industry’s characteristic high reliance on borrowed funds relative to owner’s equity—an attribute of the banking sector that, when paired with other capital and liquidity metrics, plays a crucial role in determining resilience to financial shocks.

```{r summary-stats, echo=FALSE, results='asis', fig.cap="Descriptive Statistics of Key Variables."}
# Define the variables you want to summarize
key_vars_summary <- c("esg_score", "environmental_pillar_score", "social_pillar_score",
                      "governance_pillar_score", "cet1_risk_exposure", "tier1_risk_exposure",
                      "totalcap_risk_exposure", "leverage_ratio", "provisions",
                      "provisions_ratio", "liquidity_ratio", "rwa_ratio",
                      "log_assets", "equity_to_assets") # Select relevant vars

# Generate and print the table
summary_table <- generate_summary_table(data_analysis,
                                        vars = key_vars_summary,
                                        title = "Descriptive Statistics of Key Variables")
print(summary_table)
```

## Correlations

Table \@ref(tab:correlation-matrix) displays the correlation patterns among key variables, revealing several notable relationships. The overall ESG score exhibits strong positive correlations with the social (0.83) and governance (0.85) pillars, and a moderate correlation with the environmental pillar (0.62). This suggests that while all three dimensions contribute to overall ESG ratings, governance and social factors may carry somewhat greater weight in determining overall ESG performance in European banks.

Interestingly, ESG scores show negative correlations with risk metrics: specifically, -0.25 with CET1 risk exposure. This negative relationship, though modest, aligns with the hypothesis that stronger ESG performance may be associated with lower risk levels. However, the direction of causality cannot be inferred from these correlations alone.

The three ESG pillars show different patterns of association with risk metrics. The environmental pillar correlates negatively (-0.11) with CET1 risk exposure, while the social pillar shows a stronger negative relationship (-0.35). Governance displays the weakest correlation with CET1 risk (-0.12). These varying relationships underscore the importance of examining ESG components separately rather than relying solely on aggregate scores.
Bank size (log assets) demonstrates positive correlations with ESG score (0.50) and environmental pillar score (0.58), suggesting that larger banks tend to have stronger overall ESG performance, particularly in environmental areas. This pattern might reflect larger institutions' greater resources for sustainability initiatives or their enhanced regulatory scrutiny and reporting requirements.

The leverage ratio exhibits strong positive correlations with equity-to-assets (0.87) and RWA ratio (0.86), confirming the interconnected nature of these capital adequacy measures. Notably, the leverage ratio shows a negative correlation with log assets (-0.45), indicating that larger banks tend to operate with relatively lower leverage ratios.

```{r correlation-matrix, echo=FALSE, results='asis', fig.cap="Correlation Matrix."}
# Define variables for the correlation matrix (can be a subset for readability)
cor_vars <- c("esg_score", "environmental_pillar_score", "social_pillar_score",
              "governance_pillar_score", "cet1_risk_exposure", "leverage_ratio",
              "provisions_ratio", "liquidity_ratio", "rwa_ratio",
              "log_assets", "equity_to_assets")

# Generate and print the correlation table
correlation_table <- generate_correlation_table(data_analysis,
                                                vars = cor_vars,
                                                title = "Correlation Matrix")
print(correlation_table)
```

## Variable Distributions

Figure \@ref(fig:distributions) illustrates the distributions of five key variables. The ESG score distribution appears approximately normal with a slight left skew, showing a concentration of values between 70 and 85. This clustering suggests a convergence toward relatively high ESG performance among European banks, though the left tail indicates a subset of institutions with substantially lower scores.

The CET1 risk exposure distribution exhibits a more symmetrical bell shape centered around 0.6, with most values falling between 0.5 and 0.7. This relative uniformity likely reflects the regulatory framework's influence, as banks must maintain minimum capital requirements while facing competitive and shareholder pressures that discourage excessive capital holdings.

The leverage ratio distribution shows right skew, with most observations concentrated between 0.15 and 0.25, and a longer tail extending toward higher values. This pattern indicates that while most banks maintain similar leverage levels, some institutions operate with significantly higher ratios, potentially reflecting different business models or risk appetites.

Log assets display a bimodal distribution, with peaks around 12.5 and 14.5, suggesting two distinct size clusters in our sample—medium-sized banks and larger institutions. This bimodality is important to consider in subsequent analyses, as size may moderate the relationship between ESG performance and risk metrics.

The equity-to-assets ratio distribution is right-skewed with most values concentrated between 0.04 and 0.08. The clustering at relatively low values is consistent with banking industry norms, where thin equity margins are typical. The right tail represents banks operating with more conservative capital structures.

```{r distributions, echo=FALSE, fig.cap="Distributions of Key Variables."}
# Define variables for distribution plots
dist_vars <- c("esg_score", "cet1_risk_exposure", "leverage_ratio",
               "log_assets", "equity_to_assets") # Choose a few important ones

# Generate the plot
distribution_plot <- generate_distribution_plots(data_analysis, dist_vars)

# Print the plot to include it in the document
print(distribution_plot)
```

## Time Trends

Figure \@ref(fig:time-trends) depicts the evolution of four key variables from 2020 to 2023, revealing important temporal patterns. The ESG score shows an overall positive trajectory, rising from approximately 73.1 in 2020 to 73.9 in 2023, with a temporary decline to 72.5 in 2021. This pattern suggests that despite potential pandemic-related setbacks in 2021, European banks have generally strengthened their sustainability practices over the observed period.
CET1 risk exposure displays a contrasting trend, declining from 0.67 in 2020 to approximately 0.60 in 2021-2022, before increasing slightly to 0.61 in 2023. This initial decrease might reflect regulatory flexibility during the pandemic or strategic shifts in banks' risk-weighted asset compositions. The subsequent stabilization and slight uptick suggest a potential return to pre-pandemic risk levels.
The leverage ratio exhibits more volatility, increasing from 0.21 in 2020 to 0.24 in 2021, declining to 0.23 in 2022, then rising again to 0.24 in 2023. This fluctuation may reflect banks' adaptive responses to changing economic conditions and regulatory expectations throughout the post-pandemic recovery period.
Liquidity ratio shows the most dramatic shift, rising sharply from 0.12 in 2020 to approximately 0.16 in 2021-2022, before declining to 0.14 in 2023. The initial increase likely represents a prudential response to pandemic-induced uncertainty, with banks building liquidity buffers. The subsequent decline might indicate a gradual return to more normalized liquidity management as economic conditions stabilized.

Collectively, these temporal trends reveal a complex adaptive landscape where banks have generally improved their ESG performance while navigating significant macroeconomic challenges that affected their risk and capital metrics. The non-uniform patterns across different variables highlight the multifaceted nature of banks' strategic responses to changing regulatory, economic, and sustainability pressures during this period.

```{r time-trends, echo=FALSE, fig.cap="Average Trends Over Time."}
# Define variables for time trend plots
trend_vars <- c("esg_score", "cet1_risk_exposure", "leverage_ratio", "liquidity_ratio")

# Generate the plot
time_trend_plot <- generate_time_trend_plots(data_analysis, trend_vars)

# Print the plot
print(time_trend_plot)
```


# Methodology

This section presents the empirical approach used to examine the relationship between ESG performance and banking risk. I employ several complementary methodologies to address potential issues of unobserved heterogeneity, non-linearity, and endogeneity, providing a comprehensive analysis of how ESG factors relate to various banking risk measures.

## Fixed Effects Panel Models

The primary analytical framework consists of panel data models with two-way fixed effects. This approach controls for both time-invariant bank characteristics and common time trends, addressing concerns about unobserved heterogeneity highlighted in previous banking studies (Gehrig et al., 2023; Palmieri et al., 2023). The baseline specification for the overall ESG score is:

$$Y_{i,t} = \beta_0 + \beta_1 \text{ESG}_{i,t} + \boldsymbol{\beta_2} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$

where $Y_{i,t}$ represents the risk measure for bank $i$ in year $t$ (e.g., CET1 risk exposure, Tier 1 risk exposure, leverage ratio); $\text{ESG}_{i,t}$ denotes the overall ESG score; $\mathbf{X}_{i,t}$ is a vector of control variables including the natural logarithm of total assets and the equity-to-assets ratio; $\alpha_i$ are bank fixed effects capturing time-invariant bank characteristics; $\lambda_t$ are year fixed effects accounting for common time trends; and $\varepsilon_{i,t}$ is the error term.

For a more granular analysis of ESG components, I also estimate models using the three pillar scores separately:

$$Y_{i,t} = \beta_0 + \beta_1 \text{ENV}_{i,t} + \beta_2 \text{SOC}_{i,t} + \beta_3 \text{GOV}_{i,t} + \boldsymbol{\beta_4} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$

where $\text{ENV}_{i,t}$, $\text{SOC}_{i,t}$, and $\text{GOV}_{i,t}$ represent the environmental, social, and governance pillar scores, respectively.

This differentiated approach follows Bannier et al. (2021) and Palmieri et al. (2023a), who demonstrated that individual ESG components may influence risk metrics differently. The fixed effects specification helps mitigate omitted variable bias from time-invariant factors, a crucial concern noted by Carnevale and Drago (2024) in ESG-banking research.

## Non-linear Relationships and Dynamic Specifications

To account for potential non-linear relationships between ESG performance and risk measures, as suggested by Barth et al. (2022) and Ersoy et al. (2022), I extend the baseline model to include quadratic terms:

$$Y_{i,t} = \beta_0 + \beta_1 \text{ESG}_{i,t} + \beta_2 \text{ESG}_{i,t}^2 + \beta_3 Y_{i,t-1} + \boldsymbol{\beta_4} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$

where $\text{ESG}_{i,t}^2$ is the squared term of the ESG score and $Y_{i,t-1}$ represents a lagged dependent variable to account for potential persistence in risk measures. A similar quadratic specification is applied to the pillar-level model:

$$Y_{i,t} = \beta_0 + \beta_1 \text{ENV}_{i,t} + \beta_2 \text{ENV}_{i,t}^2 + \beta_3 \text{SOC}_{i,t} + \beta_4 \text{SOC}_{i,t}^2 + \beta_5 \text{GOV}_{i,t} + \beta_6 \text{GOV}_{i,t}^2 + \beta_7 Y_{i,t-1} + \boldsymbol{\beta_8} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$

The inclusion of lagged dependent variables addresses potential dynamics in the relationship between ESG performance and risk, as banks may adjust their risk profiles gradually in response to changes in ESG factors. This dynamic specification aligns with approaches used by Danisman and Tarazi (2023) and Gehrig et al. (2023) in similar contexts.

## Quantile Regression Analysis

To examine potential heterogeneous effects of ESG factors across different levels of banking risk, I employ quantile regression analysis, following methodological precedents in banking literature such as Curcio et al. (2023) and Izcan and Bektas (2022). Unlike standard OLS or fixed effects models that focus on the conditional mean, quantile regressions estimate the relationship at different points in the conditional distribution of the dependent variable:

$$Q_{Y_{i,t}}(\tau|\text{ESG}_{i,t}, \mathbf{X}_{i,t}) = \beta_0(\tau) + \beta_1(\tau) \text{ESG}_{i,t} + \boldsymbol{\beta_2}(\tau) \mathbf{X}_{i,t}$$

where $Q_{Y_{i,t}}(\tau|\text{ESG}_{i,t}, \mathbf{X}_{i,t})$ denotes the $\tau$-th conditional quantile of the risk measure $Y_{i,t}$, and $\beta_j(\tau)$ are the coefficients that may vary across quantiles. I estimate this model for five quantiles ($\tau = 0.1, 0.25, 0.5, 0.75, 0.9$), providing insights into how ESG scores affect banks with different risk profiles. This approach is valuable for regulatory purposes, as it helps identify whether ESG factors have stronger implications for higher-risk institutions.

For pillar-specific effects, I estimate:

$$Q_{Y_{i,t}}(\tau|\text{ENV}_{i,t}, \text{SOC}_{i,t}, \text{GOV}_{i,t}, \mathbf{X}_{i,t}) = \beta_0(\tau) + \beta_1(\tau) \text{ENV}_{i,t} + \beta_2(\tau) \text{SOC}_{i,t} + \beta_3(\tau) \text{GOV}_{i,t} + \boldsymbol{\beta_4}(\tau) \mathbf{X}_{i,t}$$

This allows for examining whether specific ESG components have differential impacts across the risk distribution, a nuance that means-based approaches might obscure.

## Principal Component Analysis

To address the high dimensionality and potential multicollinearity among ESG components, I employ Principal Component Analysis (PCA), a dimensionality reduction technique. This approach is particularly valuable given the detailed ESG component data available in the Refinitiv database, which includes numerous subcategories beyond the three main pillars.

PCA transforms the original ESG variables into a set of orthogonal principal components that capture the maximum variance in the data:

$$PC_j = \sum_{k=1}^p a_{jk} Z_k$$

where $PC_j$ is the $j$-th principal component, $Z_k$ represents the standardized $k$-th ESG variable, and $a_{jk}$ are the component loadings. I retain enough components to explain approximately 80% of the total variance, with a maximum of five components to maintain interpretability.

These principal components are then used as regressors in the fixed effects model:

$$Y_{i,t} = \beta_0 + \sum_{j=1}^m \beta_j PC_{j,i,t} + \boldsymbol{\beta_{m+1}} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$

where $m$ is the number of retained principal components. This PCA approach follows methodological insights from Goss and Roberts (2010), who applied similar techniques to reduce dimensionality in KLD social responsibility data. It allows for capturing the underlying structure of ESG performance while avoiding collinearity issues that might arise when including multiple correlated ESG components simultaneously.

## Instrumental Variable Approach

To address potential endogeneity concerns arising from reverse causality or omitted variables, I implement an instrumental variable (IV) approach. Following strategies employed by Danisman and Tarazi (2023) and Gehrig et al. (2023), I construct instruments based on leave-one-out country-year averages of ESG scores:

$$\text{Instr\_ESG}_{i,c,t} = \frac{\sum_{j \in c, j \neq i} \text{ESG}_{j,t}}{\sum_{j \in c, j \neq i} 1}$$

where $\text{Instr\_ESG}_{i,c,t}$ is the instrument for bank $i$ in country $c$ in year $t$, calculated as the average ESG score of all other banks in the same country and year. This instrument leverages the idea that a bank's ESG performance is influenced by country-level norms and regulations (Ioannou and Serafeim, 2012), but should not directly affect that specific bank's risk measures except through its impact on the bank's own ESG performance.

The IV estimation is implemented using a two-stage approach:

First stage:
$$\text{ESG}_{i,t} = \gamma_0 + \gamma_1 \text{Instr\_ESG}_{i,c,t} + \boldsymbol{\gamma_2} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \nu_{i,t}$$

Second stage:
$$Y_{i,t} = \beta_0 + \beta_1 \widehat{\text{ESG}}_{i,t} + \boldsymbol{\beta_2} \mathbf{X}_{i,t} + \alpha_i + \lambda_t + \varepsilon_{i,t}$$

where $\widehat{\text{ESG}}_{i,t}$ represents the predicted values from the first stage. Similar instruments are constructed for each pillar score in the pillar-specific IV models. This approach helps isolate the causal effect of ESG performance on banking risk, addressing concerns raised by Bouslah et al. (2013) and Cornett et al. (2016) about the direction of causality between ESG and financial outcomes.

## Robustness and Specification Tests

For each model, I conduct a series of robustness and specification tests to ensure the validity of the results. For the fixed effects models, I implement Hausman tests to confirm the appropriateness of fixed versus random effects specifications. I also employ Driscoll-Kraay standard errors to address potential cross-sectional dependence, a common concern in panel data with financial institutions, as highlighted by Curcio et al. (2024).

For the IV estimations, I evaluate instrument strength using first-stage F-statistics, with values above 10 generally indicating sufficiently strong instruments (Staiger and Stock, 1997). While formal Hausman tests for endogeneity are limited by data constraints, I examine the differences between fixed effects and IV estimates to assess the potential magnitude of endogeneity bias.

In the quantile regression framework, I use bootstrap methods (Koenker and Hallock, 2001) to compute standard errors robust to heteroskedasticity. For PCA, I examine loading patterns and variance explained to ensure that the retained components adequately capture the underlying ESG dimensions.

This comprehensive methodological approach, combining fixed effects, non-linear specifications, quantile regression, PCA, and IV estimation, allows for a robust examination of the relationship between ESG performance and banking risk, addressing multiple econometric challenges highlighted in the literature. By implementing these complementary methodologies, I aim to provide a nuanced understanding of how different aspects of ESG performance relate to various banking risk measures across the risk distribution.



# Results


# Conclusion
  
# Appendix

## References

## Tables

```{r variables-table, echo=FALSE}
```{r variables-table, echo=FALSE}
source("variables_df.R")
kable(variables_df, 
      caption = "Overview of variables", 
      booktabs = TRUE,
      longtable = TRUE) %>%
  kable_styling(latex_options = c("striped", "repeat_header", "hold_position"),
                font_size = 9) %>%
  column_spec(1, bold = TRUE, width = "2.5in") %>%
  column_spec(2, width = "2.5in") %>%
  column_spec(3, width = "1in")
```

```{r}
# Display the tables
ms_esg_table
ms_pillar_table
```


```{r}
# Run the quantile regression analysis
results <- run_simplified_esg_analysis(
  data = data_analysis,  
  dep_vars = c("cet1_risk_exposure", "tier1_risk_exposure", 
               "leverage_ratio", "provisions_ratio"),
  tau_values = c(0.1, 0.25, 0.5, 0.75, 0.9)
)
```

```{r}
results$plots$combined$esg
```

\newpage

```{r}
results$plots$combined$environmental
```

```{r}
results$plots$combined$social
```

\newpage

```{r}
results$plots$combined$governance
```

```{r}
# Create summary tables
summary_data <- get_raw_summary_tables(
  results$coefficients$esg, 
  results$coefficients$pillars
)

# Display summary tables
make_simple_esg_summary_table(summary_data)
make_simple_pillar_summary_table(summary_data)

# Create detailed tables for specific variables
make_esg_table(results$coefficients$esg, "cet1_risk_exposure")
make_pillar_table(results$coefficients$pillars, "leverage_ratio")
```

```{r}
# Run the PCA analysis
pca_results <- run_pca_esg_analysis(
  data = data_analysis,  
  dep_vars = c("cet1_risk_exposure", "tier1_risk_exposure", 
               "totalcap_risk_exposure", "leverage_ratio", "provisions", 
               "provisions_ratio", "liquidity_ratio", "rwa_ratio"),
  variance_threshold = 0.8
)

# Display visualizations in the report
pca_results$visualizations$scree_plot
```

\newpage

```{r}
pca_results$visualizations$loading_plot
pca_results$visualizations$pc_interpretation
pca_results$visualizations$pc_model_table
pca_results$visualizations$comparison_table
```


```{r iv-analysis-setup, include=FALSE}
# Source the IV analysis script
invisible(source(here::here("R", "analysis_iv.R")))
```

# Instrumental Variable Analysis


```{r iv-display, fig.height=5, fig.width=8, results='hide'}
# Run the IV analysis on your dataset
iv_results <- run_iv_analysis(data_analysis)
# Make sure all visualization objects are available (hide the TRUE output)
invisible(create_iv_visualizations())
```

```{r iv-summary-table, fig.width=6}
# Format the summary table to fit better
iv_summary_table %>%
  kable_styling(font_size = 8, latex_options = "scale_down")
```

```{r iv-plots, fig.height=9, fig.width=8}
# Arrange plots in a grid with appropriate sizing
grid.arrange(
  iv_comparison_plot + theme(plot.margin = margin(0.1, 0.1, 0.5, 0.1, "cm")),
  iv_pct_diff_plot + theme(plot.margin = margin(0.5, 0.1, 0.1, 0.1, "cm")),
  ncol = 1,
  heights = c(1.3, 1)
)
```

```{r}
iv_pillar_table
```

