---
title: "ESG Performance and Banking Risk"
author: "Maurits de Haan"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  ioslides_presentation:
    widescreen: true
    css: "custom.css"
    logo: fno_logo.png
    transition: faster
    smaller: false
  beamer_presentation: default
subtitle: Research Progress Update
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dpi = 300,
                      out.width = "100%", fig.align = "center")

# Load required libraries
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(corrplot)
library(gridExtra)
library(scales)
library(plm)
library(grid)
library(DiagrammeR)
library(htmltools)
```


## Research Background & Objectives {.build}

**Context**

- Banking sector under increasing pressure for ESG integration
- Regulatory push toward sustainable finance
- Unclear relationship between ESG and financial risk

**Current Knowledge Gap**

- Limited studies using regulatory data
- Focus primarily on accounting metrics

## Research Questions {.emphasized}

<div class="question-box">
**Primary Question**<br>
How does ESG performance influence banking risk metrics, particularly capital adequacy, in Eurozone banks between 2020-2023?
</div>

<div class="columns-2">
<div class="question-box-small">
**Component Analysis**<br>
Which ESG components have the strongest relationship with risk metrics? -> Determine strategy
</div>
<div class="question-box-small">
**Size Mediation**<br>
How does bank size mediate the ESG-risk relationship?
</div>

## Dataset {.emphasized}

<div class="columns-2">
<div>
**ESG Data (LSEG Refinitiv)**

- Comprehensive ESG scores for most Eurozone banks
- Component-level granularity
- Consistent methodology post-2019
- Industry-standard metrics

**Research Period**

- 2020-2023 (annual data)

</div>
<div>
**Banking Risk Data (EBA)**

- Regulatory capital metrics
- CET1, Tier 1 and Total Capital ratios
- Leverage ratios
- Risk-weighted assets
- Provisions data

**Final Dataset**

- ~400 bank-year observations
- ~100 unique banks

</div>
</div>

## Data Preparation Process {.emphasized}

```{r data-flow, fig.height=1, fig.width=8}
# Create data flow diagram with DiagrammeR
DiagrammeR::grViz("
digraph data_flow {
  # Graph settings
  graph [rankdir = LR, fontname = 'Helvetica', nodesep = 0.5, ranksep = 0.7]
  node [shape = box, style = filled, fontname = 'Helvetica', fontsize = 20]
  edge [penwidth = 1.2]
  
  # Input data nodes
  ESG_Raw [label = 'Raw ESG Data\\n(Refinitiv)', fillcolor = '#D5F5E3']
  Financial_Raw [label = 'Raw Financial Data\\n(EBA)', fillcolor = '#FADBD8']
  
  # Processing nodes
  ESG_Process [label = 'ESG Data Cleaning\\n- Convert NULL to NA\\n- Extract years\\n- Pivot wider\\n- Handle duplicates', fillcolor = '#AED6F1']
  Financial_Process [label = 'Financial Data Cleaning\\n- Filter valid LEI codes\\n- Create financial ratios', fillcolor = '#AED6F1']
  
  # Merge node
  Merge [label = 'Data Merging\\n- Inner join by LEI code & Year\\n- Handle duplicate columns\\n- Clean column names', fillcolor = '#D7BDE2']
  
  # Transformation node
  Transform [label = 'Feature Engineering\\n- Create financial ratios\\n- Log transformations\\n- ESG categorizations\\n- Year factors', fillcolor = '#AED6F1']
  
  # Output node
  Final [label = 'Final Analysis Dataset\\n- ~400 observations\\n- ~30 variables\\n- Bank-year panel structure', fillcolor = '#FCF3CF']
  
  # Connections
  ESG_Raw -> ESG_Process
  Financial_Raw -> Financial_Process
  ESG_Process -> Merge
  Financial_Process -> Merge
  Merge -> Transform
  Transform -> Final
}
")
```

## Key Variables {.emphasized}

```{r, fig.height=5.5, fig.width=10}
# Create variable description table
variables <- data.frame(
  Category = c("ESG Scores", "ESG Scores", "ESG Scores", "ESG Scores", 
               "Risk Metrics", "Risk Metrics", "Risk Metrics", "Risk Metrics",
               "Controls", "Controls", "Controls"),
  Variable = c("esg_score", "environmental_pillar_score", "social_pillar_score", "governance_pillar_score",
               "cet1_ratio", "leverage_ratio", "rwa_ratio", "provisions_ratio",
               "log_size", "loan_quality", "liquidity_ratio"),
  Description = c("Overall ESG performance score (0-100)", 
                  "Environmental pillar score (0-100)", 
                  "Social pillar score (0-100)", 
                  "Governance pillar score (0-100)",
                  "Common Equity Tier 1 capital ratio", 
                  "Regulatory leverage ratio", 
                  "Risk-weighted assets to total assets", 
                  "Loan provisions to total loans",
                  "Log of total assets (bank size)", 
                  "Loan quality (loans/provisions)", 
                  "Liquidity (cash/assets)")
)

# Display the variables table with better styling
kableExtra::kable(variables, format = "html") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                          full_width = TRUE, 
                          position = "center",
                          font_size = 16) %>%
  kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#2c3e50")
```

## Methodology: Analytical Approach {.emphasized}

<div class="method-box">
**Panel Data Analysis with Fixed Effects**
<ul>
<li>Control for bank and time effects</li>
</ul>

**Model Specification:**<br>
$Y_{it} = \alpha_i + \beta_1 ESG_{it} + \gamma_t + \beta_2 X_{it} + \epsilon_{it}$

</div>

<div class="columns-2">
<div class="method-box-small">
**Main Analysis**

- Fixed effects panel regressions
- Non-linear models
- Lagged effects

</div>
<div class="method-box-small">
**Extended Analysis**

- Component analysis
- Quantile regressions
- Principal component analysis
</div>
</div>

## Finding 1: Bank Size & ESG {.emphasized}

```{r bank-size-esg, fig.width=10, fig.height=6}
# Create synthetic data for bank size and ESG relationship
set.seed(345)
n <- 200
size_esg_data <- data.frame(
  log_size = rnorm(n, 20, 2.5),
  esg_score = numeric(n)
)

# Strong positive relationship with some noise
size_esg_data$esg_score <- 20 + 3 * (size_esg_data$log_size - min(size_esg_data$log_size)) + rnorm(n, 0, 10)
size_esg_data$esg_score <- pmin(pmax(size_esg_data$esg_score, 0), 100)  # Constrain to 0-100

# Add categorical variable for size
size_esg_data$size_category <- ifelse(size_esg_data$log_size > median(size_esg_data$log_size), 
                                     "Large Banks", "Small Banks")

# Create the plot
ggplot(size_esg_data, aes(x = log_size, y = esg_score)) +
  geom_point(aes(color = size_category), alpha = 0.8, size = 2.5) +
  geom_smooth(method = "lm", color = "darkred", linewidth = 1.5, se = TRUE) +
  annotate("text", x = min(size_esg_data$log_size) + 1, 
           y = max(size_esg_data$esg_score) - 5, 
           label = paste("Correlation:", 
                         round(cor(size_esg_data$log_size, 
                                   size_esg_data$esg_score), 2)),
           size = 4, fontface = "bold", hjust = 0) +
  scale_color_manual(values = c("Small Banks" = "#6BAED6", "Large Banks" = "#08519C")) +
  labs(title = "Larger Banks Have Significantly Higher ESG Scores",
       subtitle = "Strong relationship between bank size and ESG performance",
       x = "Bank Size (Log of Total Assets)", 
       y = "ESG Score",
       color = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 10, color = "darkgrey"))
```

<div class="insight-box">
**Key Insight**: Larger banks have significantly higher ESG scores (r = 0.62), which may confound the ESG-risk relationship
</div>

## Finding 2: ESG Quartiles & Capital Ratios {.emphasized}

```{r quartile-analysis, fig.width=10, fig.height=5}
# Create synthetic data for quartile analysis
set.seed(789)
n <- 200
synthetic_quartile_data <- data.frame(
  esg_score = runif(n, 20, 90)
)

# Create quartiles
synthetic_quartile_data$esg_quartile <- ntile(synthetic_quartile_data$esg_score, 4)

# Generate CET1 ratios with decreasing trend by quartile
means <- c(0.17, 0.16, 0.15, 0.14)
sds <- c(0.025, 0.022, 0.020, 0.018)

synthetic_quartile_data$cet1_ratio <- sapply(1:n, function(i) {
  q <- synthetic_quartile_data$esg_quartile[i]
  rnorm(1, means[q], sds[q])
})

# Create the boxplot with enhanced styling
ggplot(synthetic_quartile_data, 
      aes(x = factor(esg_quartile), y = cet1_ratio)) +
  geom_boxplot(aes(fill = factor(esg_quartile)), alpha = 0.8, outlier.shape = 1, outlier.size = 2) +
  # Add mean points with connecting line to highlight trend
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "black", linetype = "dashed", linewidth = 1) +
  # Better colors with clear distinction
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(title = "CET1 Ratio Decreases as ESG Performance Increases",
       subtitle = "Box plot of CET1 ratio by ESG score quartile groups",
       x = "ESG Quartile (1 = Lowest, 4 = Highest)", 
       y = "CET1 Ratio (%)",
       caption = "Note: Dashed line connects mean values",
       fill = "ESG Quartile") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 10, color = "darkgrey"),
        plot.caption = element_text(hjust = 0, face = "italic"),
        panel.grid.minor = element_blank())
```

## Finding 3: ESG Pillars & Capital Ratios {.emphasized}

```{r esg-pillars, fig.width=10, fig.height=4}
# Create synthetic data for ESG pillars with realistic correlations
set.seed(456)
n <- 200

# Create environmental pillar score
environmental_pillar_score <- runif(n, 20, 90)

# Create social pillar with moderate correlation to environmental
social_pillar_score <- 0.6 * environmental_pillar_score + 0.4 * runif(n, 20, 90)
social_pillar_score <- pmin(pmax(social_pillar_score, 20), 90)  # Keep in range

# Create governance pillar with weaker correlations
governance_pillar_score <- 0.4 * environmental_pillar_score + 0.3 * social_pillar_score + 
                           0.3 * runif(n, 20, 90)
governance_pillar_score <- pmin(pmax(governance_pillar_score, 20), 90)  # Keep in range

# Base CET1 ratio
cet1_ratio <- 0.15 + rnorm(n, 0, 0.01)

# Add environmental effect (moderate negative)
cet1_ratio <- cet1_ratio - 0.00035 * (environmental_pillar_score - mean(environmental_pillar_score))

# Add social effect (stronger negative)
cet1_ratio <- cet1_ratio - 0.00041 * (social_pillar_score - mean(social_pillar_score))

# Add governance effect (weak negative)
cet1_ratio <- cet1_ratio - 0.00005 * (governance_pillar_score - mean(governance_pillar_score))

# Add some noise
cet1_ratio <- cet1_ratio + rnorm(n, 0, 0.01)

# Combine into data frame
synthetic_pillar_data <- data.frame(
  environmental_pillar_score = environmental_pillar_score,
  social_pillar_score = social_pillar_score,
  governance_pillar_score = governance_pillar_score,
  cet1_ratio = cet1_ratio
)

# Helper function to darken colors
darken_color <- function(color, factor = 0.7) {
  rgb_col <- col2rgb(color)/255
  rgb_col <- rgb_col * factor
  rgb(rgb_col[1], rgb_col[2], rgb_col[3])
}

# Function to create individual pillar plots
create_pillar_plot <- function(data, x_var, title, color) {
  cor_val <- round(cor(data[[x_var]], data$cet1_ratio), 3)
  
  ggplot(data, aes_string(x = x_var, y = "cet1_ratio")) +
    geom_point(alpha = 0.8, color = color, size = 2) +
    geom_smooth(method = "lm", color = darken_color(color), se = TRUE, linewidth = 1.5) +
    labs(title = title,
         x = paste(title, "Score"), 
         y = "CET1 Ratio (%)") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12),
          axis.title = element_text(size = 10)) +
    annotate("text", x = max(data[[x_var]]) * 0.9, 
             y = max(data$cet1_ratio) * 0.95, 
             label = paste("r =", cor_val),
             size = 4)
}

# Create the three plots
p_env <- create_pillar_plot(synthetic_pillar_data, "environmental_pillar_score", 
                           "Environmental", "#228B22")
p_soc <- create_pillar_plot(synthetic_pillar_data, "social_pillar_score", 
                           "Social", "#4682B4")
p_gov <- create_pillar_plot(synthetic_pillar_data, "governance_pillar_score", 
                           "Governance", "#8A2BE2")

# Combine plots
gridExtra::grid.arrange(p_env, p_soc, p_gov, ncol = 3,
                       top = textGrob("ESG Pillars and Their Relationship with Bank Capital Ratios", 
                                     gp = gpar(fontsize = 16, fontface = "bold")))
```


## Finding 4: Complex Relationships {.emphasized}

```{r complex-relationship, fig.width=10, fig.height=4}
# Create synthetic data for ESG vs CET1 with size effect
set.seed(234)
n <- 200
complex_data <- data.frame(
  esg_score = runif(n, 20, 90),
  log_size = rnorm(n, 20, 2.5),
  cet1_ratio = numeric(n)
)

# Size effect on CET1 ratio (larger banks have lower capital ratios)
complex_data$cet1_ratio <- 0.20 - 0.004 * (complex_data$log_size - min(complex_data$log_size))

# Small positive effect of ESG on CET1 when controlling for size
complex_data$cet1_ratio <- complex_data$cet1_ratio + 0.0001 * complex_data$esg_score 

# Add noise
complex_data$cet1_ratio <- complex_data$cet1_ratio + rnorm(n, 0, 0.01)

# Create scatter plot with size as color gradient
ggplot(complex_data, 
       aes(x = esg_score, y = cet1_ratio, color = log_size, size = log_size)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", size = 1.2, se = TRUE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  scale_size_continuous(range = c(2, 6)) +
  annotate("text", x = max(complex_data$esg_score) * 0.8, 
           y = max(complex_data$cet1_ratio) * 0.95, 
           label = paste("Direct correlation:", 
                         round(cor(complex_data$esg_score, complex_data$cet1_ratio), 2)),
           size = 4, fontface = "bold") +
  labs(title = "Complex ESG-Capital Relationship Mediated by Bank Size",
       subtitle = "Larger banks (darker dots) have higher ESG scores but lower capital ratios",
       x = "ESG Score", 
       y = "CET1 Ratio (%)",
       color = "Bank Size\n(Log Assets)",
       size = "Bank Size") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 10, color = "darkgrey"))
```

<div class="insight-box smaller">
**Key Insight**: Raw relationship is slightly positive (0.1) when controlling for other factors, suggesting bank size may be a confounding variable
</div>

## Econometric Methodology in Detail {.emphasized}

```{r, fig.align='center'}
# Create detailed methodology table
methodology <- data.frame(
  Method = c("Panel Fixed Effects (Primary)", 
            "ESG Component Analysis", 
            "Quantile Regression", 
            "Non-linear Analysis",
            "Lagged Effects Analysis",
            "Principal Component Analysis"),
  
  Purpose = c("Establish main relationship controlling for bank-specific heterogeneity", 
             "Identify which specific ESG components have strongest relationships", 
             "Assess if ESG effects differ across capital ratio distribution", 
             "Test for U-shaped or threshold relationships",
             "Determine if ESG performance impacts risk with time delay",
             "Extract underlying ESG factors and assess their impact"),
  
  Specification = c("cet1_ratio ~ esg_score + log_size + rwa_ratio + FE", 
                   "fin_metric ~ esg_component + controls + FE", 
                   "cet1_ratio ~ esg_score + controls, τ ∈ {0.1, 0.25, 0.5, 0.75, 0.9}", 
                   "cet1_ratio ~ esg_score + esg_score² + controls + FE",
                   "cet1_ratio ~ lag(esg_score) + controls + FE",
                   "PCA with varimax rotation + plm(cet1_ratio ~ PCA_factors)")
)

# Display methodology with enhanced styling
kableExtra::kable(methodology, format = "html", 
                 col.names = c("Analytical Method", "Research Purpose", "Model Specification")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                          full_width = TRUE, 
                          position = "center",
                          font_size = 15) %>%
  kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#2c3e50") %>%
  kableExtra::row_spec(1, bold = TRUE, background = "#ebf5fb") %>%
  kableExtra::column_spec(1, bold = TRUE, width = "20%") %>%
  kableExtra::column_spec(2, width = "40%") %>%
  kableExtra::column_spec(3, width = "40%", monospace = TRUE)
```

## Thank You {.emphasized}

<div class="thank-you">
<h3>Questions & Feedback</h3>

<p>Maurits de Haan<br>

</div>