---
title: "ESG Performance and Banking Risk"
author: "Maurits de Haan"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  ioslides_presentation: 
    widescreen: true
    css: "custom.css"
    logo: fno_logo.png
    transition: faster
    smaller: false
subtitle: Data Science Lab Final Presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                     error = FALSE, fig.width = 10, fig.height = 5.5)

# Load required libraries
library(plm)
library(dplyr)
library(stargazer)
library(kableExtra)
library(modelsummary)
library(broom)
library(ggplot2)
library(gridExtra)
library(plm)
library(AER)
library(gridExtra)
library(kableExtra)
library(modelsummary)
library(plm)
library(dplyr)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(patchwork)
library(modelsummary) 
library(broom)        
library(knitr)
library(kableExtra)

# Silently run the source files to get the data and functions
sink(file = tempfile())
invisible(source(here::here("R", "import_ESG.R")))
invisible(source(here::here("R", "import_financial.R")))
invisible(source(here::here("R", "merging_ESG_financial.R")))
invisible(source(here::here("R", "analysis_fixed_effects_quadratic.R")))
invisible(source(here::here("R", "analysis_quantile.R")))
invisible(source(here::here("R", "analysis_pca.R")))
invisible(source(here::here("R", "analysis_iv.R")))
invisible(source(here::here("R", "analysis_exp.R")))
sink()

# Run the analyses that will be used in the presentation
results <- run_simplified_esg_analysis(
  data = data_analysis,  
  dep_vars = c("cet1_risk_exposure", "tier1_risk_exposure", 
               "leverage_ratio", "provisions_ratio"),
  tau_values = c(0.1, 0.25, 0.5, 0.75, 0.9)
)

pca_results <- run_pca_esg_analysis(
  data = data_analysis,  
  dep_vars = c("cet1_risk_exposure", "tier1_risk_exposure", 
               "totalcap_risk_exposure", "leverage_ratio", "provisions", 
               "provisions_ratio", "liquidity_ratio", "rwa_ratio"),
  variance_threshold = 0.8
)

iv_results <- run_iv_analysis(data_analysis)
invisible(create_iv_visualizations())
```

## Research Context

### Research Question

<div class="columns-2">
<div>
- **How does ESG performance relate to banking risk?**
  - Focus on regulatory capital adequacy metrics
  - European banks, 2020-2023
  - Multiple methodological approaches

- **Why it matters:**
  - Regulatory perspective
  - Scientific contribution
  - ESG integration in financial stability
</div>

<div>
```{r, fig.height=3.2}
time_trend_plot <- generate_time_trend_plots(data_analysis, 
                                       c("esg_score", "cet1_risk_exposure"))
print(time_trend_plot)
```
</div>
</div>

## Data Overview

### Key Variables & Descriptive Statistics

<div class="columns-2">
<div>
**ESG Variables:**
- Overall ESG score (mean: 73.3)
- Environmental pillar (mean: 86.0)
- Social pillar (mean: 75.4)
- Governance pillar (mean: 68.6)

**Key Bank Risk Metrics:**
- CET1 risk exposure (mean: 0.60)
- Leverage ratio (mean: 0.24)
- Provisions ratio (mean: 0.05)
</div>

<div>
```{r, fig.height=3.5}
distribution_plot <- generate_distribution_plots(data_analysis, 
                                        c("esg_score", "cet1_risk_exposure"))
print(distribution_plot)
```
</div>
</div>

## Methodological Framework

### Multi-Method Approach

<div class="columns-2">
<div>
1. **Fixed Effects Panel Models**
   - Controls for bank/time unobserved heterogeneity
   - Linear & non-linear specifications

2. **Quantile Regression**
   - Heterogeneous effects across risk distribution

3. **Instrumental Variables (IV)**
   - Addresses endogeneity concerns
   - Country-year averages as instruments

4. **Principal Component Analysis (PCA)**
   - Dimensionality reduction of ESG components
   - Captures underlying ESG structure
</div>

<div>
**Why multiple methods?**
- Each addresses different methodological challenges
- Triangulation for robust inference
- Complementary insights across risk distribution
- Enhanced ability to detect complex relationships

**Data constraints:**
- 132 bank-year observations
- Short time dimension (4 years)
- Need for statistically efficient approaches
</div>
</div>

## Fixed Effects Results

### ESG Effects on Capital Ratios

```{r}
# Create a streamlined version of the fixed effects table for the presentation
# Focus on key results only
fe_table <- ms_esg_table

# Display a more compact version for the presentation
kable(fe_table[1:4, 1:4], format = "html", 
      caption = "Fixed Effects Panel Models: Impact of ESG on Risk Measures")
```

- **Key Finding:** Modest relationship between overall ESG and capital ratios
- Weak positive but statistically insignificant coefficients
- Linear models may mask heterogeneous effects
- Need for more nuanced analytical approaches

## ESG Pillar-Level Findings

```{r}
# Create a streamlined version of the pillar table for the presentation
pillar_table <- ms_pillar_table[1:8, 1:4]

# Display a more compact version for the presentation
kable(pillar_table, format = "html",
      caption = "Pillar-Level Effects on CET1 Risk Exposure")
```

- **Social pillar:** Strongest positive association with capital adequacy (0.027**)
- **Environmental pillar:** Mixed effects depending on risk measure
- **Governance pillar:** Generally weak relationships
- Evidence of non-linear relationships (quadratic terms significant)

## Quantile Regression Insights

### Heterogeneous ESG Effects Across Risk Distribution

```{r, fig.height=5}
# Display the quantile regression plot for ESG score
results$plots$combined$esg
```

- **Key insight:** ESG effects vary substantially across risk distribution
- Stronger negative effect for banks with lower capital ratios
- Statistical significance diminishes for high-quantile banks
- Standard mean-based models mask this heterogeneity

## Social Pillar Quantile Effects

```{r, fig.height=5}
# Display the quantile regression plot for Social pillar
results$plots$combined$social
```

- **Social pillar:** Consistently negative effect on capital risk
- Strongest at middle-to-upper quantiles (0.50-0.75)
- Capital buffer complementarity with social responsibility
- Suggests strategic differentiation in ESG focus across risk profiles

## PCA Results: ESG Dimensionality

```{r, fig.height=5}
# Display the PCA scree plot
pca_results$visualizations$scree_plot
```

- First 5 components explain ~83% of ESG variance
- PC1 (46.2%): General ESG quality (inverse measure)
- PC2 (13.8%): Social/product responsibility vs environmental focus
- PC3 (10.2%): Stakeholder engagement vs governance
- Reveals complex multidimensional nature of ESG performance

## PCA Component Structure

```{r, fig.height=5.5}
# Display the PCA loading plot
pca_results$visualizations$loading_plot
```

- **Key insight:** ESG dimensions extend beyond traditional E-S-G pillars
- Cross-cutting components suggest strategic trade-offs
- Evidence of potential ESG dimensionality not captured by aggregate scores
- More nuanced structure relevant for understanding risk relationships

## IV vs Fixed Effects: Endogeneity

```{r, fig.height=5}
# Display the IV vs FE comparison plot
grid.arrange(
  iv_comparison_plot + theme(plot.margin = margin(0.1, 0.1, 0.5, 0.1, "cm")),
  ncol = 1
)
```

- **IV estimates substantially larger** than fixed effects
- Suggests negative bias in conventional estimates
- Evidence that endogeneity concerns are significant
- Different directions across risk metrics (e.g., RWA ratio)

## Model Performance Enhancement

```{r}
# Display the PCA vs standard model comparison
pca_results$visualizations$comparison_table %>%
  kable(format = "html", caption = "Model Performance: Original vs. PCA") %>%
  kable_styling(font_size = 14)
```

- **PCA models outperform original specifications**
- Modest improvements for CET1 risk exposure (0.003)
- Larger gains for leverage ratio (0.027) and provisions
- Evidence that dimensionality reduction enhances explanatory power
- Captures risk-relevant ESG structure more effectively

## Key Data Science Insights

### Methodological Implications

- **Complex relationships require multiple approaches**
  - No single method provides complete picture
  - Triangulation yields more robust inference

- **Quantile regression reveals heterogeneity**
  - Mean-based models mask differential effects
  - Distribution-wide perspective essential

- **Dimensionality reduction improves modeling**
  - Enhanced explanatory power
  - Reveals underlying ESG structure beyond traditional pillars

- **Endogeneity matters substantively**
  - Conventional estimates likely biased
  - Direction of causality requires careful identification

## Conclusion & Future Directions

### Toward Better ESG-Risk Modeling

- **ESG-risk relationship is nuanced:**
  - Varies across risk distribution
  - Different across ESG dimensions
  - Non-linear patterns present

- **Methodological recommendations:**
  - Employ multiple complementary approaches
  - Consider distributional heterogeneity
  - Address potential endogeneity
  - Look beyond aggregate scores

- **Extensions for future research:**
  - Longer time series with structural breaks
  - Machine learning approaches
  - Regulatory stress testing incorporation
  - Dynamic panel methods

## Thank You & Questions

```{r, fig.height=5}
# Create a summary visualization for the final slide
grid.arrange(
  results$plots$combined$esg,
  pca_results$visualizations$loading_plot,
  ncol = 2
)
```
